---
title: "Introduction to Linear Regression"
author: "Bret Larget"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning = FALSE,
                      cache=TRUE, autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(scales)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

## Example: Age and Height

- The following graph shows a plot of height in inches versus age in months of a boy from age 2 years to 8 years.

```{r}
riley = read_table("../../data/riley.txt")
df = riley %>% 
  filter(age >=2*12 & age <=8*12)
```

```{r}
ggplot(df, aes(x = age, y = height)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm")
```

## Regression Model

$$
Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i
$$

- $\varepsilon_i$ is the random "error" between the actual value $Y_i$ and the mean value $\beta_0 + \beta_1 X_i$.
- A common method to estimate $\beta_0$ and $\beta_1$ from data $x$ and $y$ is the method of "least squares"
    - This line minimizes the sum of squared vertical distances between the points and a line
    - Find values $b_0 = \hat{\beta}_0$ and $b_1 = \hat{\beta}_1$ to minimize

$$
\sum_{i=1}^n \big(y_i - (b_0 + b_1 x_i)\big)^2
$$

- There is a multi-variate calculus solution to this problem, but we explore a different way to do it.

## lm()

- We can use `lm()` to fit the linear regression model

```{r}
df_lm = lm(height ~ age, data = df)
cf = coef(df_lm)
cf
summary(df_lm)
```

- The coefficients are the slope, $b_1 = `r round(cf[2],3)`$, and the intercept $b_0 = `r round(cf[1],3)`$.
    - The slope has units inches per month (y over x)
    - The intercept has units inches (y)
- For this data set, $x=0$ is outside the range of the data
    - The value at $x=0$ has a meaningful interpretation, the height at birth
    - However, this interpretation requires an *extrapolation* beyond the range of the data which assumes that the linear relationship continues
    
### Coefficients and Summary Statistics

- The least squares coefficients have simple equations based on summary statistics of the data.

$$
b_1 = r \times \frac{s_y}{s_x}
$$

$$
b_0 = \bar{y} - b_1 \bar{x}
$$

- Note that the regression line goes through the point $(\bar{x},\bar{y})$.

```{r}
x = df$age
y = df$height

xbar = mean(x)
ybar = mean(y)
sx = sd(x)
sy = sd(y)
r = cor(x,y)

c(xbar, ybar, sx, sy, r)

b1 = r *sy/sx
b0 = ybar - b1*xbar

c(b0, b1)
cf

```

### Predicted values

- The straightforward way to make a prediction is just to plug into the regression equation with the estimated coefficients.

- Estimate the boy's height at age 50 months.

```{r}
## one way
est = b0 + 50*b1
est

## using coef()
sum(cf * c(1,50))
```

- The regression line crosses has the value $\hat{y} = 42.775$ at $x=50$ months.

#### Understand by standard units

- We can also understand this with standard units.

- The value $x=50$ is $z = (50 - \bar{x})/s_x)$ standard deviations from the mean.

```{r}
x0 = 50
z = (x0 - mean(x))/sd(x)
z
```

- We predict that $y$ will be $rz$ standard deviations from $\bar{y}$.
    - Here, $r$ is very close to one, so we predict that $y$ is nearly the same number of standard deviations from $\bar{y}$ as $x$ was from $\bar{x}$.
    
```{r}
yhat = mean(y) + r*z*sd(y)
yhat
```

#### Lion Data

- Let's try this with the lions data

```{r}
lions = read_csv("../../data/lions.csv") %>% 
  rename(black = proportion.black)

ggplot(lions, aes(x = age, y = black)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm")

lions_lm = lm(black ~ age, data = lions)
lions_cf = coef(lions_lm)
lions_cf
summary(lions_lm)
```

#### Summary Statistics

```{r}
lions_sum = lions %>% 
  summarize(across(everything(), list(mean = mean, sd = sd)),
            n = n(),
            r = cor(age, black)) %>% 
  relocate(n)

lions_sum = lions_sum %>% 
  mutate(b1 = r*black_sd/age_sd,
         b0 = black_mean - b1*age_mean)

lions_sum %>% 
  print(width = Inf)

```

- Predict at age 10 years

```{r}
yhat_1 = sum( c(1,10) * lions_cf )
yhat_1
```

#### Standard Units

```{r}
x0 = 10
z = (x0 - lions_sum$age_mean)/lions_sum$age_sd
z
```

- 10 years is 2.126 standard deviations above the mean

```{r}
lions_sum$r*z
```

- Predict the proportion black of the nose to be $rz = 1.68$ standard deviations above the mean.

```{r}
yhat_2 = lions_sum$black_mean + lions_sum$r*z*lions_sum$black_sd
yhat_2
```



```{r}

```


