


---
title: "Exoplanet Mass-Radius Relationship and Regression"
author: "Bret Larget"
output: html_document
---

This file includes many contributions from Professor Jessi Kehe.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning = FALSE,
                      cache=TRUE, autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(scales)
library(modelr)
library(tidymodels)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

### Setup details

- This lecture use the following scripts, assumed to be in your course scripts directory.
    - `COURSE/scripts/viridis.R`
    - `COURSE/scripts/ggprob.R`
    
- You also need the exoplanet data set
    - `COURSE/data/exoplanets_default_2021-06-15.csv`
    - `COURSE/data/chicago-2018.csv`

- You will need the packages:
    - **tidyverse**
    - **scales**
    - **tidymodels**
    - **modelr**







```{r}
## Read in the csv file
## Select confirmed planets, rename some variables
planets_orig = read_csv("../../data/exoplanets_default_2021-06-15.csv",
                   skip = 100)

planets = planets_orig %>% 
  select(pl_name, hostname, discoverymethod, disc_year,
         sy_pnum, pl_rade, pl_bmasse) %>%
  rename(planet = pl_name,
         star = hostname,
         method = discoverymethod,
         year = disc_year,
         number = sy_pnum,
         radius = pl_rade,
         mass = pl_bmasse)
```



## Regression Overview


### Background

- Many problems in Statistics and Data Science involve examining the relationship between variables.
- A common setting is when:
    - a single quantitative variable is deemed a *response variable*
    - one or more other variables are treated as *explanatory variables* or *predictors*.
- Such models are referred to as *linear models*
    - In a standard linear model, the response variable is assumed to have a normal distribution given the predictors.

#### Special cases

- Linear models can be categorized based on the number and types of the predictors.
    - *simple linear regression*: a single quantitative predictor
    - *multiple regression*: multiple quantitative predictors
    - *one-way analysis of variance*: a single categorical predictor
    - *analysis of variance*: multiple categorical predictors

#### Generalizations

- When the distribution of the response variable is not modeled with a normal distribution, such as when it is a count or a single 0/1 response, it is often preferable to use a *generalized linear model*.
    - *logistic regression*: for 0/1 response variables;
    - *Poisson regression*: for non-negative count data.
- At other times, it is preferable to model a categorical predictor as having a distribution itself.
Some of these models are called *mixed effects models*.

## Example

- International Arrival airport wait time data from Chicago in 2018.

```{r data-chicago}
chicago = read_csv2("../../data/chicago-2018.csv")
chicago
```

- Model the number of booths as a function of the total number of passengers.
    - How did the airport plan on staffing?
    
```{r graph-chicago}
chicago_1 = chicago %>%
  select(all_total,all_booths) %>%
  drop_na()

ggplot(chicago_1, aes(x=all_total,y=all_booths)) +
  geom_point(alpha=0.1) +
  geom_smooth(method = "lm")

fit_1 = lm(all_booths ~ all_total, data=chicago_1) #lm(response_var ~ explanatory variable)

summary(fit_1)

cf = coef(fit_1)

ggplot(chicago_1, aes(x=all_total,y=all_booths)) +
  geom_point(alpha=0.1) +
  geom_smooth(method = "lm", se=FALSE, size = 2) +
  geom_abline(slope = cf[[2]], intercept = cf[[1]], color="magenta")
```

- See that `geom_smooth(method = "lm")` adds the same regression line we found with `lm()`.
- There are about 10 booths plus an additional 7.5 per 1000 passengers arriving per hour, on average.





## Review of Correlation

The correlation coefficient, $r$, is a summary statistic describing a particular relationship between two quantitative variables.  

The correlation coefficient...

- Measures the strength and direction of a linear relationship  
- Is used for quantitative variables  
- Does not distinguish the two variables  
- Has no units of measurement  
- $-1 \leq r \leq 1$

$$
r = \frac{\sum_{i=1}^n \left(\frac{x_i - \bar{x}}{s_x}\right)
\left(\frac{y_i - \bar{y}}{s_y}\right)}{n-1}
$$


### Thought Quiz

1. Why is there no distinction between the explanatory and response variables in correlation?  
2. Why do both variables have to be quantitative?  
3. How does changing the units of measurement change correlation?  
4. Why doesn’t a tight fit to a horizontal line imply strong correlation?  









## Review of Simple Linear Regression

Correlation tells us about the strength and direction of a line.  

- What if we want a description of how the variables vary together?

The regression line...  

- A regression line is a straight line that describes how a response variable y changes as an explanatory variable x changes  
- We can use a regression line to predict the value of y for a given value of x  
- The distinction between explanatory and response variables is important  
- The Least-squares regression line is the unique line such that the sum of the squared vertical (y) distances between the data points and the line is as small as possible.

#### Regression formulas

The simple linear model has the form

$$
y = \beta_0 + \beta_1 x + \varepsilon
$$
where $y$ is the response variable, $x$ is the explantory variable, $b_0$ is the y-intercept, $b_1$ is the slope, and $\varepsilon$ is the error.  

Using observations $(x_1,y_1), \ldots, (x_n,y_n)$, we can estimate $b_0$ and $b_1$ to get our Least-squares regression line as 

$$
\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i,
$$
where $\hat{y}_i$ is the predicted response corresponding to $x_i$, and $\hat{\beta}_0$ and $\hat{\beta}_1$ are the estimated intercept and slope, respectively.

Least-squares Estimation: minimize the squared errors with respect to the unknown parameters.  

- This is accomplished by minimizing the following with respect to the unknown parameters:  
$$
\sum_{i=1}^n(y_i - \hat{y}_i)^2 = \sum_{i=1}^n(y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_i))^2
$$

- This results in the following estimators:

Slope:
$$
\hat{\beta}_1 = r \frac{s_y}{s_x}
$$
where $r$ is the correlation coefficient, $s_y$ is the sample standard deviation for $y$, and $s_x$ is the sample standard deviation for $x$.

Intercept:
$$
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}
$$
where $\bar{y}$ and $\bar{x}$ are the sample means of the $y$ and $x$ variables, respectively.

##  Regression Example

We are going to generate a fake data set to use for our example.  By using fake data, we can also check how close our regression estimates match the true values we used to simulate our data.


```{r}
## Generate our fake data set
set.seed(246810)
n = 100 ## sample size
b0 = 1  ## intercept
b1 = 2.5  ## slope

fake = tibble(
  x = runif(n, -3, 10),  ## explanatory variable
  y = rnorm(n, b0+b1*x, 3))  ## response variable

ggplot(fake, aes(x = x, y = y)) +
  geom_point()
```


```{r}
fake_sum = fake %>% 
  summarize(across(everything(),list(mean = mean, sd = sd)),
            n = n(),
            r = cor(x,y),
            slope = r*y_sd/x_sd,
            intercept = y_mean - slope*x_mean) %>% 
  relocate(n)

fake_sum %>% 
  print(width = Inf)
```

We also can use `lm()` to estimate the slope and intercept:

```{r}
lm0 = lm(y~x, data = fake)
summary(lm0)

cf = coef(lm0)
cf
```

### Residual plots

One approach for checking if our linear assumption holds is to create a residual plot.  Recall that residual corresponding to the $i$th observation is $r_i = y_i - \hat{y}_i$.

- Here we use **modelr** functions `add_residuals()` and `add_predictions()` to the fake data set.

```{r}
fake = fake %>%
  add_residuals(lm0) %>%
  add_predictions(lm0)

ggplot(fake, aes(x=x, y =resid)) +
  geom_point() +
  xlab("x") +
  ylab("Residuals") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")

```


As a comparison, below is a residual plot for a different simulated data set.

Do you see any pattern in this residual plot?

```{r, echo = FALSE}
## Generate a different fake data set that is not linear
set.seed(246810)
n = 100 ## sample size
x1 = runif(n, -3, 3)  ## explanatory variable
y1 = rnorm(n,0.5*x1^2,1)  ## response variable

df1 = tibble(x=x1, y=y1)
lm1 = lm(y~x, df1)

df1 = df1 %>%
  add_residuals(lm1) %>%
  add_predictions(lm1)

ggplot(df1, aes(x=x, y=resid)) +
  geom_point() +
  xlab("x") +
  ylab("Residuals") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")
```

There is a clear pattern.  The residuals near the boundaries are positive and the residuals near the middle of the x range are negative. 

This suggests that the relationship between y and x is not linear and appears to be quadratic.






## Overview of Exoplanet Mass-Radius Relationship

In previous lectures, we explored data from the NASA Exoplanet Archive.  Two variables that are useful for characterizing an exoplanet are mass and radius.

The two most prolific methods for detecting exoplanets are the Transit Method and the Radial Velocity Method. 

- The Transit Method works by monitoring the total light output of a star across time.  
    - If a planet crosses between the star in the line-of-sight of the telescope, we may observe a dip in the light output of the star.  
    - A dip that repeats at a regular period suggests that a planet may be present.
    - Often at least three periodic dips are necessary for the detection to be considered credible.
    - Here is a YouTube video that illustrates the Transit Method:  https://youtu.be/RrusIZaWDW8  
    - The depth of the transit (i.e., the depth of the dip in the light output) can be used to estimate the planet's radius relative to the host star.
    
- The Radial Velocity (RV) Method looks for a wobble in the star suggesting an object may be orbiting it.  
    - The host star is observed on multiple nights (often 30 or more) and a spectrum is collected using a spectrograph each nights (sometimes multiple times in a night).  
    - From the observed spectrum, some techniques are used to estimate how fast the star was moving toward or away from the observed at the time the observation was collect (i.e., the radial velocity is estimated).  
    - The estimated radial velocities are plotted against time.  If the points follow a particular pattern, it suggests the motion of the star may be due to a planet orbiting.  
    - Here is a YouTube video that illustrates the RV Method:  https://youtu.be/tUzDKlaTHFM  
    - Some properties of the shape of the fit RV curve can be used to estimate the mass of the orbiting exoplanet; more precisely, the minimum mass of the planet can be estimated.
    

### Mass-Radius Relationship

The *mass-radius relationship* of exoplanets is the relationship between exoplanets’ radius, R, their mass, M.

Modeling the relationship between mass and radius is important for the following reasons:

- Prediction  
    - The model can be used to predict a planet’s mass given its radius measurement  
    - We have more observations with radius estimates than mass estimates, so having a way to estimate mass can be useful
    
```{r}
## Number of mass and radius estimates
planets %>%
  select(mass, radius) %>%
  summarize_all(function(x) sum(!is.na(x)))
```
    
- Learning about exoplanets compositions  
    - Planet compositions can be inferred by their density  
    - Exoplanets can have a range of compositions such as rocky or gaseous  
    - Knowing about planet composition may help to understand planetary formation and evolution processes  
    - For more information about planet compositions, see [exoplanet compositions](http://astro140.courses.science.psu.edu/theme4/census-and-properties-of-exoplanets/exoplanet-composition/)
    
    
#### Plot Mass vs. Radius - quick look 

Let's take a quick look at what the Mass-Radius relationship looks like for our exoplanet data.  We'll talk later about a way to model these data.

```{r}
## How many observations do we have with both mass and radius estimates?
planets %>%
  select(mass, radius) %>%
  drop_na() %>%
  nrow()


ggplot(planets, aes(radius, mass)) +
  geom_point() +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)")
```

It's hard to see any clear pattern in this plot.  Let's adjust the axis scales to see if that helps.


```{r}
ggplot(planets, aes(radius, mass)) +
  geom_point() +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(se=FALSE) +
  geom_smooth(method="lm", se=FALSE, color="magenta")

```

Now we can see a bit more of a relationship between mass and radius on this log10 scale.

The *general* pattern is that there is a positive association between log10(radius) and log10(mass).









## Power-Law Relationship 

One of the popular models used for the Mass-Radius relationship is a power-law relationship:

$$
y = C \times x^\theta
$$
where $y$ is the response variable, $x$ is the exoplantory variable, $C$ is a scaling factor, and $\theta$ is the power law coefficient.

#### Examples of power laws

```{r}
power_law = function(theta)
{
  df = tibble(x = seq(0, 10, by = .1),
               y =x^theta)
  gg = ggplot(df, aes(x,y)) +
    geom_line() +
    ggtitle(paste0("Power law exponent: ", theta))
  return(gg)
}

power_law(1)
power_law(0.5)
power_law(2)
```


#### Power-law model for Mass-Radius relationship


While we will consider mass on the y-axis and radius on the x-axis, astronomers will often plot and model these the other way (with radius on the y-axis).  

- Since the mass measurements tend to be harder to obtain, we can look at our mass vs. radius model as useful for using an estimated radius to predict an unknown mass.


Astronomers have found that the power law relationship between mass and radius is not constant across the range of values, but instead there seems to be a different power law exponent for different mass ranges. 

- This results in what is known as a *broken power law model* where different ranges of the data have different power law parameters.  
- We will only consider a subset of the data where the power law exoponent is thought to be constant.  We define this subset below.  
- The range for masses we will consider is between 2 and 127 Earth masses.  This range comes from work by Jingjing Chen and David Kipping in 2016 where they took a data-driven approach to detect *change points* in the broken power law model.  
    - Chen, J. and Kipping, D., 2016. Probabilistic forecasting of the masses and radii of other worlds. The Astrophysical Journal, 834(1), p.17.  
    - In their model, they considered radius vs. mass, but found the noted mass range to be consistent with previous work looking for change points in radius.
    
    
Below we define the data subset we will use for our model and plot mass vs. radius.

```{r}
mr = planets %>%
  filter(between(mass, 2, 127)) %>%
  drop_na()

ggplot(mr, aes(radius, mass)) +
  geom_point() +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(se=FALSE) +
  geom_smooth(method="lm", se=FALSE, color="magenta")
```










## Fitting the Power Law Model

We already saw the form of the power law relationship is $y = C \times x^\theta$. Now we'd like to turn this into a statistical model that we can use for the exoplanet data.

The statistical model we will actually fit is going to use a log10 transformation of the power law relationship in order to make the form linear:
$$
\begin{align*}
y_i = \log10(m_i) & = \log10(C) + \theta\log10(r_i) + \varepsilon_i \\
& = \log10(C) + \theta x_i + \varepsilon_i, i = 1, \ldots, n.
\end{align*}
$$
In this model, the response variable $y_i = \log10(m_i)$ is the $\log10$(mass) for exoplanet $i$, the explanatory variable $x_i = \log10(r_i)$ is the $\log10$(radius) for exoplanet $i$, $\log10(C)$ is the (unknown) intercept, $\theta$ is the (unknown) slope, and $\varepsilon_i$ is the random error for exoplanet $i$.

- Now we have a linear model and can use methods we've already learned to fit the model!

```{r}
lm1 = lm(log10(mass) ~ log10(radius), data = mr)
summary(lm1)
```

The estimated intercept is `r round(coef(lm1)[1],3)` and the estimated slope is `r round(coef(lm1)[2],3)`.

Let's check out our fit model on a plot:

```{r}
ggplot(mr, aes(radius, mass)) +
  geom_point() +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)") +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(aes(slope = coef(lm1)[2] , intercept = coef(lm1)[1]), color="blue")
```

Let's also see what this looks like on the original scale.  

```{r}
mr %>%
  mutate(mass_pred = 10^coef(lm1)[1]*radius^coef(lm1)[2]) %>%
ggplot(aes(radius, mass)) +
  geom_point() +
  geom_line(aes(y = mass_pred), color="red") +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)")
```


Recall that the estimated slope is the power law exponent.  Since it is somewhat near to 1, this suggests that the data on the original scale are not too far from linear in radius.  

- How does our power law model compare to a simple linear regression model fit to our data on the original scale?

```{r}
mr %>%
  mutate(mass_pred = 10^coef(lm1)[1]*radius^coef(lm1)[2]) %>%
ggplot(aes(radius, mass)) +
  geom_point() +
  geom_line(aes(y = mass_pred), color="red") +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)") +
  geom_smooth(method="lm", se=FALSE, color="blue")

summary(lm(mass ~ radius, data=mr))
```

These models are quite different.  We will do some model checking on the appropriateness of our linear model on the log10 scale next.










## Model checking

The linear model we fit uses least squares regression.  

- This means that the parameters were estimated to minimize the total sum of squared errors.  
-  Let's create a plot that displays these errors.

```{r}
mr = mr %>%
  add_residuals(lm1) %>%
  add_predictions(lm1)

ggplot(mr, aes(x=radius, y=mass)) +
  geom_point() +
  geom_segment(aes(xend = radius, yend = 10^pred), color="magenta") +
  xlab("Radius (Earth Radius)") +
  ylab("Mass (Earth Mass)") +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(aes(slope = coef(lm1)[2] , intercept = coef(lm1)[1]), color="blue")
```


The vertical magenta bars are the residuals for the model defined as $r_i = y_i - \hat{y}_i$, where $\hat{y}_i$ is the predicted $\log10$(mass) from our model for exoplanet $i$.  
 
- Our fit linear model is such that the sum of the lengths of these vertical lines is minimized.  
- That is, if we drew a different line on this plot and found the errors, the sum of those errors squared would be greater than those for our `lm1` fit.

#### Residual plot

Next we are going to consider a residual plot.  This is where we remove the fit model from the data, and only plot the errors (the residuals) against radius.

```{r}
ggplot(mr, aes(x=radius, y=resid)) +
  geom_point() +
  xlab("Radius (Earth Radius)") +
  ylab("Residual (Earth Mass)") +
  scale_x_log10() +
  geom_hline(aes(yintercept = 0), color = "red", linetype = "dashed")
```


Patterns in a residual plot can suggest that our linear model model may not be appropriate for the data.  

- In this case, you may notice that the residuals corresponding to smaller radius values tend to be positive, and there seems to be a little bit of clustering of points (e.g., above 10 Earth Radius).  
- But, overall, the model form seems reasonable.









## Introduction to inference on linear models


####  Assumptions on the random errors

Let's go back to our linear model, 
$$
y_i = \log10(C) + \theta x_i + \varepsilon_i, i = 1, \ldots, n.
$$

We only mentioned that the $\varepsilon_i$'s are random errors, but we did not discuss other assumptions.

It is common to make the following assumptions (which should also be checked):  

- $E(\varepsilon_i) = 0$.  
    - This implies $E(y_i) = \log10(C) + \theta x_i$.  
- Errors have a constant variance:  Var$(\varepsilon_i) = \sigma^2$.  
    - This implies Var$(y_i) = \sigma^2$.  
- The errors are uncorrelated.

When desiring to do inference on the estimated parameters, another common assumption to make is that the errors are normally distributed, $\varepsilon_i \sim N(0, \sigma)$.  

  - This implies that $y_i \sim N(\log10(C) + \theta x_i, \sigma)$.

This normality assumption on the errors has implications for the estimates of our parameters.  In particular, it has the consequence that the estimators of our intercept and slope are normally distributed.  


## Recall: Z-scores and t-scores

In a previous discussion assignment you were introduced to Z-scores and t-scores.  The t-scores are going to show in our regression inference, so we review that information here.  

> Note that in this review, the estimator is the sample mean.  When we get back to regression the estimators will be for the slope or intercept, which will result in some changes in the t-statistic and the degrees of freedom of the resulting t-distribution.

Assume a model where $X_1,\ldots,X_n$ are drawn at random from a distribution with a mean $\mu$ and a standard deviation $\sigma$.

The sampling distribution of the sample mean,
$\bar{X} = n^{-1}\sum_{i=1}^n X_i$ has a mean $\mu$ and standard deviation $\sigma/\sqrt{n}$.

A mathematical derivation is required to show this formally, or simulation can be used to check if it the expressions are plausible.

#### Z-Score

We have seen in many settings that the z-statistic (substract the mean, divide by the standard deviation) often has an approximate standard normal distribution.
$$
Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}
$$
If the distribution of each $X_i$ is normal, then $Z$ will be normal as well.

Even if $X_i$ does not have a normal distribution, the distribution of $\bar{X}$ will be approximately normal if $n$ is large enough to overcome the non-normality, a result known as the central limit theorem.

#### t distribution

However, $\sigma$ is typically unknown and things are a bit different when the sample standard deviation $s$ is substituted for $\sigma$.
$$
T = \frac{\bar{X} - \mu}{S/\sqrt{n}}
$$
where
$$
S = \sqrt{ \frac{\sum_{i=1}^n (X_i - \bar{X})^2}{n-1} }
$$

The added random variability in the denominator means that even when the distribution of a single random variable is exactly normal, the distribution of $T$ is not.

Instead, it has a $t$ distribution with $n-1$ degrees of freedom, which is a bell-shaped density centered at zero, but more spread out than a standard normal density.

When the degrees of freedom becomes large, the $t$ distribution is quite close to standard normal.

It is identical to standard normal when the degrees of freedom is infinite.

#### R functions

These R functions are similar to their normal counterparts.

- `rt()`: generate random variables from a t distribution
- `pt()`: find an area under a t density
- `qt()`: find a quantile from a t density
- `dt()`: return the height of the t density

In addition, the following functions are available in the script ggprob.R to add t densities to plots.

- `geom_t_density()`: add a t density to a plot
- `geom_t_fill()`: add a filled t density
- `gt():` graph a t density

The following graph shows a standard normal distribution in black and t distributions with degrees of freedom equal to 1, 2, 4, 8, \ldots, 1028 ranging from yellow to violet.

```{r show-t, echo=FALSE, fig.height=4}
col = viridis(10,begin=1,end=0)
g = ggplot()
  
for ( i in 1:10 )
  g = g + geom_t_density(2^(i-1),color=col[i],a=-5,b=5)

g = g +
  geom_norm_density(color="black") +
  geom_hline(yintercept=0) +
  theme_bw()

plot(g)
```

#### Confidence Intervals

A confidence interval for $\mu$ has the form
$$
\bar{x} \pm t^* \frac{s}{\sqrt{n}}
$$
where $t^*$ is selected so that the area between
$-t^*$ and $t^*$ under a t density with $n-1$ degrees of freedom is the desired confidence level.

A confidence interval for a difference between means,
$\mu_1 - \mu_2$,
has the form
$$
\bar{x}_1 - \bar{x}_2 \pm t^* 
  \sqrt{ \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} }
$$
where $t^*$ is from a $t$ distribution
where the degrees of freedom is estimated as a function of the sample sizes and standard deviations.
Use the function `t.test()`.
This approach assumes that the standard deviations of the two populations need not be the same.

#### Hypothesis Tests

When using t distribution methods,
p-values are found by calculating the t statistic and finding areas under t distributions.


## Back to regression

These concepts will be used to carryout inference on the estimated parameters of our simple linear model, which we will do next.









## Inference for linear models

If we make the assumption that the errors on linear model are normally distributed, we can carryout hypothesis test on our estimated parameters.

We will focus on inference for the slope parameter since that tends to be the more scientifically interesting parameter.  

The hypothesis test we will carryout is
$$
H_0:  \theta = 1 \\
H_a:  \theta \neq 1
$$

We test the null that our slope parameter $\theta$ (which is the power law exponent) is 1, suggesting a linear relationship between $\log10$(mass) and $\log10$(radius), against the alternative that there is a power law relationship.

This leads to the test statistic
$$
T = \frac{\hat{\theta} - 1}{s_{\hat{\theta}}}
$$
where $s_{\hat{\theta}}$ is the standard error of $\hat{\theta}$.  Note that the 1 is from the null hypothesis assumption that $\theta = 1$.

- In case you were curious, the formula for this standard error is
$$
s_{\hat{\theta}} =  \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2/(n-2)}{\sum_{i = 1}^n (x_i - \overline{x})^2}}
$$
where $\overline{x}$ is the sample mean of the $x_i$, and $\hat{y}_i$ is the predicted $\log10$(mass) values for observation $i$.  

- We can calculate this as follows:
```{r}
n = nrow(mr)
syy = sum(mr$resid^2)
sxx = sum((log10(mr$radius) - mean(log10(mr$radius)))^2)
sqrt(syy/(n-2)/sxx)  ## standard error using the formula above
coef(summary(lm1))[2, "Std. Error"] ## standard error from our lm1 model
```
This is also the standard error that would be used in a confidence interval for $\theta$:  $\hat{\theta} \pm t_{n-2} s_{\hat{\theta}}$, where $t_{n-2}$ is selected based on the desired confidence level.  

- You may be wondering why we have a $t$ distribution with $n-2$ degrees of freedom here.  This is because our model has two parameters (the slope and intercept) so we "use up" two degrees of freedom when estimating them. 


The output of our estimated model gives us the result of (a different) hypothesis test:
$$
H_0:  \theta = 0 \\
H_a:  \theta \neq 0
$$

```{r}
summary(lm1)
```

The estimated slope is $\hat{\theta} = 1.09067$ with an estimated standard error of $SE(\hat{\theta})=0.05451$.

This leads to a t-statistic of
$$
T = \frac{1.09067 - 0}{0.05451} = 20.01
$$

We see that there are 313 observations in our data set.  Since there are two estimated parameters in our linear model (slope and intercept), we use 313-2 = 311 degrees of freedom for the t-distribution for our slope.
```{r}
## Number of observations
mr %>%
  nrow()

## Compute our p-value
pt(20.01, df=311, lower.tail=FALSE)*2 ## P(T >= 20.01) x 2
```

Notice that this p-value is very small leading us to reject the null hypothesis that the slope is zero.




To carryout our hypothesis test with
$$
H_0:  \theta = 1 \\
H_a:  \theta \neq 1
$$
we use the following t-statistic:
$$
T = \frac{1.09067 - 1}{0.05451} = 1.663457
$$

```{r}
## t-statistic
tstat = (coef(summary(lm1))[2]-1)/coef(summary(lm1))[2, "Std. Error"]
tstat

## Compute our p-value
pt(tstat, df=311, lower.tail=FALSE)*2 ## P(T >= 1.663) x 2
```

Since the p-value is greater than 0.05, there is not evidence suggesting that the power law model is appropriate.  In other words, the results are consistent with a linear model being appropriate.


We can create a plot to visualize the p-value:

```{r}
gt(311) +
  geom_vline(aes(xintercept = c(-abs(tstat), tstat)), color="red", linetype="dashed") +
  geom_t_fill(311, a = abs(tstat), b=6) +
  geom_t_fill(311, a=-6, b = -abs(tstat))

```










## Inference for linear models via simulation

Instead of relying on the theoretical values for the standard deviation of the parameters, we could instead run a simulation.

We can use the parametric bootstrap, which involves generating many realizations of the data using the initial estimate for $\theta$ and then fitting the regression model on the simulated data set to obtain many estimates of $\theta$.  The mean and standard deviation of these values can be used for inference.

Here are the steps for the parametric bootstrap:

1. Estimate the parameters of the statistical model.
2. Use the estimated model and simulate a new data set of the same size.
    - In a regression framework, this would involve adding to each predicted value $y^*$ a new simulated normally distributed random error from a distribution with mean zero and the estimated standard deviation.
3. Fit a linear model to the simulated data set and estimate the coefficients.
4. Repeat steps 2 and 3 many times.
5. Calculate the standard deviations of the coefficients from the simulated data sets as estimates of the standard errors.

```{r simulation-parametric-bootstrap}
n = mr %>%
  select(mass, radius) %>%
  drop_na() %>%
  nrow()
sigma_resid = sd(mr$resid)
N = 10000
b_0 = numeric(N)
b_1 = numeric(N)

for ( i in 1:N )
{
  mr_new = mr %>%
    drop_na() %>%
    mutate(mass = 10^(pred + rnorm(n,0,sigma_resid))) # these are the "new" y's
  lm2 = lm(log10(mass) ~ log10(radius), data=mr_new)
  b_0[i] = coef(lm2)[1]
  b_1[i] = coef(lm2)[2]
}

df_coef = tibble(b_0,b_1)
```



Let's look at the slope parameter.  We can estimate the mean and standard deviation from the parametric bootstrap simulation, and then compare it to a normal distribution with the same mean and standard deviation.

```{r}
mean_slope = mean(df_coef$b_1)
mean_slope
sd_slope = sd(df_coef$b_1)
sd_slope


ggplot(df_coef, aes(x=b_1)) +
  geom_density() +
  xlab("theta") +
  ylab("Density") +
  geom_norm_density(mu = mean_slope, sigma = sd_slope, color="blue") +
  ggtitle("Exoplanet Mass-Radius Relationship",
          subtitle = "Parametric bootstrap distribution of the slope (black), and a normal density (blue)")
```

The simulated distribution (black curve) follows the normal distribution (blue curve) quite well.

So how do these values compare to what we calculated with the theoretical equations?

```{r}
paste("Bootstrap: ", round(mean_slope,3), round(sd_slope,3))
paste("Theoretical: ", round(coef(lm1)[2],3), round(coef(summary(lm1))[2, "Std. Error"],3))
```

Very close as well!





