---
title: "NCAA Men's Division 1 Basketball Results"
author: "Bret Larget"
date: "4/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE, cache=FALSE)
library(tidyverse)
library(lubridate)
library(stringr)
library(rvest)
```

## Load and Parse the Data

I figured out how to get only the division 1 games (no opponents from other divisions).
The web says there are 350 division 1 teams in 32 conferences plus 3 teams in transition that are playing in division 1 now.
The data I have has 353 teams, so I think it is right.

I am using package `rvest` with functions to read the html which has the data formatted in a single large fixed-format file, then use `read_table()` to read this in, followed by a lot of `dplyr` and `tidyr` stuff to transform the data.

There are 353 teams and a total of 5328 games among the teams.

```{r}
## URL for data
## Data is formatted in fixed width format as a giant character
## sub=11590 is only NCAA division 1 games
url_head = "https://www.masseyratings.com/"
url_body = "scores.php?s=309912&sub=11590&all=1&mode=2&sch=on&format=0"
url = str_c(url_head, url_body)

## Read the html
ncaa_html = read_html(url)
## Parse
games = ncaa_html %>% 
  html_nodes("pre") %>% 
  html_text() %>% 
## transform  
  read_table(col_names = FALSE) %>% 
  dplyr::rename(date = X1,
         winner = X2,
         w_score = X3,
         loser = X4,
         l_score = X5,
         overtime = X6,
         neutral_site = X7) %>% 
## get rid of last row which is all NAs  
  filter(!is.na(w_score)) %>% 
## get rid of @s, but keep track of home team in case wanted later  
  mutate(h_winner = str_detect(winner,"@"),
         h_loser = str_detect(loser,"@")) %>% 
  mutate(winner = str_remove(winner,"@"),
         loser = str_remove(loser,"@")) %>% 
## change blank values in overtime into NAs  
  mutate(overtime = case_when(
    overtime == "" ~ NA_character_,
    TRUE ~ overtime))

nrow(games)

rm(url, url_body, url_head, ncaa_html)
```

## Summarize the Data

```{r}
## Get the record of each team
teams = games %>% 
  select(winner, loser) %>% 
  dplyr::rename(W = winner,
         L = loser) %>% 
  pivot_longer(cols = everything(), names_to = "result", values_to = "team") %>% 
  group_by(team,result) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = result, values_from = n) %>% 
  select(team, W, L)

nrow(teams)
```
