---
title: "Normal Distributions"
author: "Bret Larget"
output: html_document
---

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}

This R Markdown document includes contributions from Professor Jessi Kehe

### Setup details

- You will need the package **tidyverse** for this file

- This lecture use the following scripts, assumed to be in your course scripts directory.
    - `COURSE/scripts/viridis.R`
    - `COURSE/scripts/ggprob.R`

- The file `ggprob.R` contains a number of functions for graphing probability distributions in a tidyverse-friendly manner.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(egg)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```


## Normal Distributions

## Overview

- Normal distributions are continuous distributions

- All normal distributions have the same overall shape:  symmetric, unimodal, bell-shaped

- The shape of the normal density curve is completely determined by its mean ($\mu$) and its standard deviation ($\sigma$):

- The density function has the following formula

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$$

- The mean $\mu$ is located at the center and the standard deviation $\sigma$ controls the spread.

```{r normal-density}
mu = 3
sigma = 0.5
mu2 = 7
sigma2 = 2

gnorm(mu=mu, sigma=sigma, a=-0, b=12, color="blue",
      fill=NULL, title=FALSE) +
  geom_norm_density(mu = mu2, sigma = sigma2, color = "magenta") +
  theme_minimal()
```


## Normal Probabilities

- Suppose that $X \sim \text{N}(\mu, \sigma)$
    - this means the random variable $X$ follows a normal distribution with mean $\mu$ and standard deviation $\sigma$.

- We can calculate probabilities for a Normal distribution using R:

$$P(a \leq X \leq b)$$

- This is the area under the density curve between $b$ and $a$
- The function `pnorm()` calculates areas under normal curves.

- The arguments in order are: 
    - `q`: a location on the x axis
    - `mean`: the mean of the distribution, or $\mu$, default value is 0
    - `sd`: the standard deviation of the distribution, or $\sigma$, default value is 1
    - `lower.tail`: logical to calculate area to the left, default value is `TRUE`
    - `log.p`: logical to calculate log of the probability, default value is `FALSE`.
    
### Examples:

$X 
#### Area to the left

- Find $\prob(X < 80)$

```{r}
## naming arguments; I usually do not
pnorm(q=80, mean=100, sd=25)
## without the names
pnorm(80, 100, 25)

## gnorm is defined in ggprob.R
## you need to call source("../../scripts/ggprob.R") in set-up for this to work
## a is the left endpoint of the interval to fill
## b is the right endpoint of the interval to fill
## use NULL for the default value (all the way to the left or right)
gnorm(mu = 100, sigma = 25) +
  geom_norm_fill(mu = 100, sigma=25, a = NULL, b = 80) +
  theme_minimal()
```

#### Area to the right

- Find $\prob(X > 90)$

```{r}
## using lower.tail
pnorm(90, 100, 25, lower.tail = FALSE)
## using subtraction
1 - pnorm(90, 100, 25)

## first argumment is mu, second is sigma if you do not want to specify names
## Need to repeat mu and sigma in each level: not set in aes()
gnorm(100, 25) +
  geom_norm_fill(100, 25, a = 90, b = NULL) +
  theme_minimal()
```

#### Area between two values

- Find $\prob(50 < X < 120)$
    - Note: $\prob(50 < X < 120) = \prob(X < 120) - \prob(X < 50)$

```{r}
## use subtraction
pnorm(120, 100, 25) - pnorm(50, 100, 25)

gnorm(100, 25) +
  geom_norm_fill(100, 25, a = 50, b = 120) +
  theme_minimal()
```

#### Two tail areas

- Find $\prob(|X-100| > 30)$
    - This is $\prob(X < 70) + \prob(X > 130)$

```{r}
## using lower.tail
pnorm(70, 100, 25) + pnorm(130, 100, 25, lower.tail = FALSE)
## using subtraction
pnorm(70, 100, 25) + (1 - pnorm(130, 100, 25))

gnorm(100, 25) +
  geom_norm_fill(100, 25, a = NULL, b = 70) +
  geom_norm_fill(100, 25, a = 130, b = NULL) +
  theme_minimal()
```

## Standard Normal Distribution and Standardization

- A *standard normal distribution* is a normal distribution with mean = 0 and standard deviation = 1

```{r}
gnorm() +
  theme_minimal()
```


-  If $X \sim \text{N}(\mu, \sigma)$, then $Z = \left(\frac{X-\mu}{\sigma}\right) \sim \text{N}(0, 1)$

- An implication is $\prob(X < a) = \prob\left( Z < \frac{a-\mu}{\sigma}\right)$

#### Example

$X \sim \text{N}(100, 25)$

- Find $\prob(X < 80)$

```{r}
mu = 100
sigma = 25
x = 80
z = (x-mu)/25

## Using X directly
pnorm(x, mu, sigma)

## Standardizing first
## default values for the mean and sd are 0 and 1
pnorm(z)

gnorm(mu, sigma) +
  geom_norm_fill(mu, sigma, b=x) +
  theme_minimal()

gnorm() +
  geom_norm_fill(b=z) +
  theme_minimal()
```

- Notice the graphs are identical except for axis labels
- All normal curves have exactly the same shape, but they may differ in center and scale

## Normal Quantiles

- The p-quantile is the location `q` where the area to the left of `q` under the normal density is equal to `p`.

- The R function `qnorm()` makes these calculations.
- Arguments in order are:
    - `p`: the probability of the quantile
    - `mean`: mean of the distribution, $\mu$, default is 0
    - `sd`: standard deviation of the distribution, $\sigma$, default is 1
    - `lower.tail`: default is `TRUE`, if `FALSE` find locations for areas to the right
        - I **never** use this.
    - `log.p`: default is `FALSE`, if true, it returns natural log of the quantile
    
### Examples

$X \sim \text{N}(100, 25)$

Find the 0.1, 0.25, 0.9, 0.95, 0.975, and 0.99 quantiles of the distribution.

```{r}
p = c(0.1, 0.25, 0.9, 0.95, 0.975, 0.99)
mu = 100
sigma = 25

qnorm(p, mu, sigma)
```

### Graphs

```{r, fig.height = 12}
## use ggarrange() to display multiple plots and map() to create a list of plots

normal_plots = p %>% 
  map(~{
    gnorm(mu, sigma) +
    geom_norm_fill(mu, sigma, a = NULL, b = qnorm(.x, mu, sigma)) +
    theme_minimal()
  })

ggarrange(plots = normal_plots,
          nrow = length(p),
          ncol = 1)
```

### Simple Normal Calculation example

> The weights of packets of cookies produced by a certain manufacturer have a normal distribution with a mean of 202 g and a standard deviation of 3 g. What is the weight that should be labeled on the packet so that only 1% of the packets are underweight?

- $X \sim \text{N}(202, 3)$, want $P(X < a) = 0.01$  

```{r simple-normal}
## This gives the 1st percentile of a N(202, 3) distribution
qnorm(0.01, mean=202, sd=3, lower.tail=TRUE)

## Check
pnorm(195.021, mean=202, sd=3) # P(X <= 195.021) = .01
```


## Normal approximation to the binomial distribution

- If $n$ is large enough, and $p$ is not too close to 0 or 1, the binomial distribution can be approximated by the normal distribution 
- A rule of thumb for "large enough" is if $np(1-p) \ge 10$.
    - Under the assumptions above, if $X \sim \text{Binomial}(n,p)$, then $X \stackrel{\text{approx}}{\sim}\text{N}(np, \sqrt{np(1-p)})$
    

- Here is an example where the assumption is satisfied.
    - $n = 100$
    - $p = 0.5$
    - $np(1-p) = 25$
    
```{r normal-approx-good}
## Assumptions satisfied
n = 100
p = 0.5
mu = n*p
sigma = sqrt(n*p*(1-p))

gbinom(n, p, scale=TRUE) +
  geom_norm_density(mu, sigma, color = "red") +
  theme_minimal()
```

- Examine the accuracy by comparing $P(X \le k)$
    - Exact calculation uses `pbinom(k, n, p)`
    - Normal approximation uses `pnorm(k+0.5, mu, sigma)`
    
- Note the use of `k+0.5` instead of `k` for the approximation
- The height of the binomial probability at `k` may be represented by the area of a rectangle with the same height between `k - 0.5` and `k + 0.5`.

- `dbinom(k, n, p)` is approximately equal to `pnorm(k+0.5, mu, sigma) - pnorm(k, mu, sigma)`

```{r normal-approx-good-graphs}
### check the accuracy of normal approximation 
df = tibble(k = 0:n,
            binomial_tail = pbinom(k, n, p),
            binomial_prob = dbinom(k, n, p),
            normal_tail = pnorm(k+0.5, mu, sigma),
            normal_prob = pnorm(k+0.5, mu, sigma) - pnorm(k-0.5, mu, sigma))

## Accuracy of left probabilities
ggplot(df, aes(x = binomial_tail, y = normal_tail - binomial_tail)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(label = scales::comma) +
  xlab("Exact Binomial Probability") +
  ylab("Error in Normal Approximation") +
  ggtitle("Errors in Normal Approximation to Binomial",
          subtitle = "left tail probabilities") +
  theme_minimal()

## Accuracy of individual probabilities
ggplot(df, aes(x = binomial_prob, y = normal_prob - binomial_prob)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(label = scales::comma) +
  xlab("Exact Binomial Probability") +
  ylab("Error in Normal Approximation") +
  ggtitle("Errors in Normal Approximation to Binomial",
          subtitle = "individual outcome probabilities") +
  theme_minimal()
```

### Example where the approximation is poor

- Here is an example where the assumption is satisfied.
    - $n = 100$
    - $p = 0.01$
    - $np(1-p) = 0.99$
    
```{r normal-approx-bad}
## Criteria not satisfied
n = 100
p = 0.01
mu = n*p
sigma = sqrt(n*p*(1-p))

gbinom(n, p, scale=TRUE) +
  geom_norm_density(mu, sigma, color = "red") +
  theme_minimal()
```

- Error analysis

```{r normal-approx-bad-graphs}
### check the accuracy of normal approximation 
df = tibble(k = 0:n,
            binomial_tail = pbinom(k, n, p),
            binomial_prob = dbinom(k, n, p),
            normal_tail = pnorm(k+0.5, mu, sigma),
            normal_prob = pnorm(k+0.5, mu, sigma) - pnorm(k-0.5, mu, sigma))

## Accuracy of left probabilities
ggplot(df, aes(x = binomial_tail, y = normal_tail - binomial_tail)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(label = scales::comma) +
  xlab("Exact Binomial Probability") +
  ylab("Error in Normal Approximation") +
  ggtitle("Errors in Normal Approximation to Binomial",
          subtitle = "left tail probabilities") +
  theme_minimal()

## Accuracy of individual probabilities
ggplot(df, aes(x = binomial_prob, y = normal_prob - binomial_prob)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(label = scales::comma) +
  xlab("Exact Binomial Probability") +
  ylab("Error in Normal Approximation") +
  ggtitle("Errors in Normal Approximation to Binomial",
          subtitle = "individual outcome probabilities") +
  theme_minimal()
```






