---
title: "Volleyball and Modeling"
output:
  html_document: default
---

\newcommand{\given}{\,|\,}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning = FALSE,
                      cache=TRUE, autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(scales)
library(tidymodels)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

```{r default-theme, include=FALSE, eval=TRUE}
## reset default theme
theme_set(theme_bw())
```

### Background

- Volleyball is a popular sport around the world.
- The traditional version is played indoors with teams of six players on the court.
- Beach volleyball is a popular variant with two-person teams and is played on sand.
- Both versions of volleyball are Olympic sports and there are professional leagues of both styles around the world.
- In the US, both boys and girls play sanctioned high school volleyball and there is a strong club competition too.
- At the collegiate level, women's volleyball is regulated by the National Collegiate Athletic Association (NCAA).
- Teams compete in three separate divisions, with Division I being the most competitive.
- There are 340 Division I teams during the Fall 2021 season.

#### Wisconsin

- The UW Volleyball team won the 2021 Big 10 volleyball championship.
- They are the #4 overall seed in a 64-team single elimination tournament.
- The [full bracket](https://www.ncaa.com/brackets/volleyball-women/d1/2021) is available on the NCAA web site.
- The Badgers beat Colgate and Florida Gulf State University in straight sets last weekend to advance to the Sweet 16.
- They swept UCLA is straight sets yesterday and advance to play Saturday evening against Minnesota, whom they have beaten twice this year during the regular season, sweeping them in Minnesota on October 1 and prevailing in five sets here in Madison the weekend before Thanksgiving.
- All of the Badgers NCAA tournament games have been played here at home.
- A victory on Saturday would take them to the Final Four next weekend in Columbus, Ohio.
- Go Badgers!

#### Rules

- The important volleyball rules for us to know is that each match consists of sets and each set is series of rallies where the teams take turns hitting a volleyball across a net to the other team until the rally ends with one team winning the rally.
- There are various rules we do not need to model about legal hits and various violations, but the winner of each rally scores one point.
- The first team to win three sets wins the match.
- The first four sets (the fourth is only played if needed) are won by the first team to reach 25 points with a lead of at least two points.
    - If the score of a match is tied at 24 points each, play continues until a team has a lead of two points at which point that team wins the set.
- Should a fifth set be needed (if each team wins two of the first four sets), the rules are the same, but the target is 15 points instead of 25.

#### Modeling Preview

- We will model each volleyball match as a series of Binomial trials with the same probability that a given team wins the point versus the other team.
- For a given match, this probability will be determined by the relative strengths of each teams, these unknown strengths being parameters of interest we will estimate from the data of all regular season matches.

### Data

- Data for all NCAA Division I volleyball matches is found on [the NCAA website](https://stats.ncaa.org/season_divisions/17720/scoreboards?utf8=%E2%9C%93&season_division_id=&game_date=12%2F09%2F2021&conference_id=0&tournament_id=&commit=Submit)
- There is a separate URL to produce a table with results for each date.
- I used the **rvest** package to scrape this data from the web by:
    - looping over the dates
    - reading each webpage
    - creating a list where each element was a row
    - processing this data to extract important data
- The code is messy as what looks like a nice table on the web page has the information from a single match spread over five different rows with different numbers of elements, so it is not as easy as just grabbing the whole table with a single command.
- The data requires some more processing
    - We eliminate preseason exhibition games
    - We eliminate matches played against non-Division I opponents
    - Some team names need to be modified if they include rank numbers
    - We join in conference information from another file from the NCAA.
- This processed data is saved in the file made available to you.
- There are 24 matches with data anomalies, such as a set with a final score of 22-16.
- Fixing these errors would require finding for each match an alternative source of data.
- The overall effect of mis-recording a few points from fewer than half of a percent of all matches is not likely to influence our analysis.

### 2021 Women's NCAA Volleyball Data

- The code block below reads in this data and eliminates matches from the national tournaments.

```{r}
vb_orig = read_csv("../../data/vb-division1-2021-all-matches-corrected.csv")

vb = vb_orig %>% 
  filter(is.na(site) |
           (!str_detect(site, "National Volleyball Invitational Championship") &
               !str_detect(site, "NCAA DI Volleyball Championship")))

n_matches = nrow(vb)

teams_orig = read_csv("../../data/vb-division1-2021-teams.csv")

teams = teams_orig %>% 
  select(Team) %>% 
  mutate(team_index = row_number()) %>% 
  relocate(team_index)

conferences = teams_orig %>% 
  select(Team, Conference) %>% 
  mutate(team_index = row_number()) %>% 
  relocate(team_index)

n_teams = nrow(teams)
```

- The data includes  the scores for each set across the nearly 5,000 individual matches between Division I opponents during the 2021 women's volleyball season.
- The (mostly) cleaned data set has `r n_matches` matches involving only these `r n_teams` division I teams from the regular season or conference tournaments.
    - I do not include results from the NCAA Division I tournament or the National Volleyball Invitational tournament.
    - There are about two dozen matches where the final scores in sets are not consistent with the rules of volleyball
    - The final winning scores for some sets have something other than 25 (or more than 25 and winning by exactly 2) in the first four sets or winning by something other than 15 (or more than 15 and winning by exactly 2) in the fifth set
    - The discrepancies are just a few points in the rare cases they happen
    - The overall impact on the inference is likely quite small.

## Model

I pose a model where the strength of each team is represented by a parameter $\theta$. The difference between the values of this parameter for two opponents (possibly with adjustments for home court and random effects) determine the chance that a team wins a point. Assume that a match is made up of a series of independent Bernoulli trials, each resulting in a point for one team or the other, until the match is won.
This model assumes the chance of a team winning a point is the same whether or not the team is serving and assumes that the chance does not depend on the current score of the game.
These assumptions cannot be checked using only data from the scores of each set from each match, but are reasonable.

The simple model for the probability that team $i$ with strength $\theta_i$ wins a single point versus team_2 with strength $\theta_j$ is a function of
$$
\Delta_{ij} = \theta_i - \theta_j
$$
where the probability is
$$
\mathsf{P}(\Delta_{ij}) = \frac{1}{1 + \mathrm{e}^{-\Delta_{ij}}}
$$
Note that this probability is equal to
$\frac{\mathrm{e}^{\Delta_{ij}}}{1 + \mathrm{e}^{\Delta_{ij}}}$
and that the probability that team $j$ wins the point is one minus this probability, or $\frac{1}{1 + \mathrm{e}^{\Delta_{ij}}}$.

For more complicated models that also depend on a home-court advantage and a random match effect, replace $\Delta_{ij}$ with $\Delta_{ij} + \alpha_{ijk} + \alpha_k$ where $\alpha_{ijk}$ is equal to $\alpha$ if team $i$ is home in match $k$, $-\alpha$ is team $j$ is home during match $k$ and 0 if match $k$ is at a neutral site.
The match random effect $\alpha_k$ is assumed to be drawn from a normal mean zero distribution.
The simple model is an example of a *logistic regression model*
and the complicated model is an example of a *mixed effects logistic regression model*.

Notice that the model requires a constraint on the values of $\theta_i$ because adding or subtracting the same value to all $\theta$ does not change the point probabilities.
In a maximum likelihood setting, this can be accomplished by restricting the sum (or mean) of the $\theta_i$ values to be fixed.
A Bayesian approach might put a prior density on the values with mean zero and an unknown standard deviation.

#### The Inverse-Logistic Function

The logistic function is a function of the log odds, $\ln(p/(1-p))$,
where $p$ is a probability.
The inverse of this function is how we model the probability:
$p = \frac{1}{1 + \mathrm{e}^{-\Delta_{ij}}}$.

```{r logistic-curve}
inv_logistic = function(x) { return ( 1/(1 + exp(-x)) )}
delta = seq(-5,5,length.out=1001)
p = inv_logistic(delta)
df = tibble(delta,p)

ggplot(df, aes(x=delta,y=p)) +
  geom_line(color="blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 1, color="red", linetype = "dashed")
```

### Likelihood

Here are results of one volleyball match from the 2019 tournament.

```{r likelihood example, echo=FALSE}
library(kableExtra)
match_1 = tibble(
  teams = c("Northeastern", "Green Bay"),
  `1` = c(16,25),
  `2` = c(28,26),
  `3` = c(25,27),
  `4` = c(23,25),
  sets = c(1,3))

total_1 = 16+28+25+23
total_2 = 25+26+27+25

kable(match_1) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE)

```

This match had a total of `r total_1 + total_2` points of which team 1, Northeastern, won `r total_1` and team 2, Green Bay, won `r total_2`.
The likelihood of this result is
$$
L(\Delta) = \mathsf{P}(\Delta)^{92}(1 - \mathsf{P}(\Delta))^{103}
$$
which achieves its maximum value when $\mathsf{P}(\Delta) = \frac{92}{195}$,
or
$$
\Delta = \ln \left( \frac{92/195}{103/195}\right) = 
\ln \left( \frac{92}{103}\right) \doteq `r round(log(92/103),4)`
$$

Estimation with many matches is much more complicated,
even in the simple model,
because the same teams play multiple games against different opponents.

### Model Summary

- Data
    - $x_k$ is the total points of team 1 in match $k$, for $k=1,\ldots,n$ (this is the sum of their points across all sets)
    - $y_k$ is the total points of team 2 in match $k$, for $k=1,\ldots,n$ (again, summed over all sets)
- Parameters
    - $\theta_i$ is the strength of team $i$, for $i=1,\ldots,`r n_teams`$
        - constrained so that $\sum_i \theta_i = 0$
- Point Probability
    - When team $i$ plays team $j$, the probability that team $i$ wins a point is a function of $\Delta_{ij} = \theta_i - \theta_j$
    - This probability is $p(\Delta) = 1/(1 + \exp(-\Delta))$

### Log-Likelihood

- The log-likelihood is the sum of the likelihoods over the match

$$
\ln L = \sum_{i=1}^k (x_k \ln p_k + y_k \ln (1-p_k))
$$
where
$$
p_k = \frac{1}{1 + \exp(\theta_{i[k]} - \theta_{j[k]})}
$$
where $\theta_{i[k]}$ and $\theta_{j[k]}$ are the strengths of team 1 and team 2 in the $k$th match.

### Data Set-up

- To wrangle the data in order to fit the model, we need to add index variables for each team.
- We also want to replace the scores over sets with the total number of points that each team scores.
- This block of code accomplishes this.

```{r data}
total_points = vb %>%
  mutate(x = s1_1 + s1_2 + s1_3 + 
           if_else(is.na(s1_4), 0, s1_4) +
           if_else(is.na(s1_5), 0, s1_5),
         y = s2_1 + s2_2 + s2_3 + 
           if_else(is.na(s2_4), 0, s2_4) +
           if_else(is.na(s2_5), 0, s2_5),
         home = as.integer(is.na(site))) %>%
  left_join(teams, by = c("team1" = "Team")) %>% 
  rename(index_1 = team_index) %>% 
  left_join(teams, by = c("team2" = "Team")) %>% 
  rename(index_2 = team_index) %>% 
  select(index, index_1, team1, conference1, index_2, team2, conference2, x, y, home)  
```

### Maximum Likelihood

- This model includes a strength parameter $\theta_i$ for each team constrained so that $\sum_{i=1}^n \theta_i = 0$ and a single parameter $\alpha$ which is the home court advantage for a team if there is one team is the home team.
- This code uses the function `optim()` to maximize the parameters.

```{r mle}
n_teams = nrow(teams)
theta_0 = rep(0, n_teams-1)
alpha = 0

inv_logistic = function(x) { return ( 1/(1 + exp(-x)) ) }

calc_logl = function(theta,alpha,df)
{
  delta = theta[df$index_1] - theta[df$index_2] - alpha*df$home
  p = inv_logistic(delta)
  return ( sum( df$x*log(p) + df$y*log(1-p)) )
}

f = function(alpha_theta,df)
{
  alpha = alpha_theta[1]
  theta = c(alpha_theta[-1],-sum(alpha_theta[-1]))
  return ( calc_logl(theta,alpha,df) )
}

theta_0 = rep(0,n_teams)
out = optim(theta_0,f,method="BFGS",
            control=list(fnscale=-1),df=total_points)

alpha = out$par[1]
theta = c(out$par[-1],-sum(out$par[-1]))
teams = teams %>%
  mutate(theta = theta) %>% 
  left_join(conferences) %>% 
  relocate(Conference, .after = Team)
```

### Explore

```{r explore}
ggplot(teams, aes(x=theta)) +
  geom_density() +
  geom_hline(yintercept = 0) +
  ggtitle("2021 NCAA Women's Volleyball Team Strength")

ranked_teams = teams %>%
  arrange(desc(theta)) %>%
  mutate(rank = row_number()) %>%
  relocate(rank, team_index)
```


```{r}
ranked_teams %>%
  group_by(Conference) %>%
  summarize(mean_theta = mean(theta)) %>%
  arrange(desc(mean_theta))

ranked_teams %>% print(n=Inf)

ranked_teams %>%
  filter(Team %in% c("Wisconsin", "Colgate", "FGCU","UCLA","Minnesota"))
```

### Compare Results with the Data

```{r}

```

### Simulation

- We next explore simulation to calculate the probability of winning a match
- For any two teams, the model produces an estimate of the probability that each team wins a point
- By simulation, we determine the probability that each team wins the match.

```{r simulation}
inv_logistic = function(x) { return ( 1/(1 + exp(-x)) )}

## simulate a single set to the target score of 25 or 15
sim_set = function(delta,target)
{
  p = inv_logistic(delta)
  points = c(0,0)
  repeat
  {
    pt = rbinom(1,1,p)
    if ( pt == 1 )
      points[1] = points[1] + 1
    else
      points[2] = points[2] + 1
    if ( max(points) >= target && abs(diff(points)) >= 2 )
      break
  }
  return ( points )
}

## simulate a match by simulating the sets
sim_match = function(delta)
{
  tab = matrix(NA,2,5)
  sets = c(0,0)
  index = 1
  repeat
  {
    if ( sum(sets) < 4 )
      result = sim_set(delta,25)
    else
      result = sim_set(delta,15)
    if ( result[1] > result[2] )
      sets[1] = sets[1] + 1
    else
      sets[2] = sets[2] + 1
    tab[,index] = result
    index = index + 1
    if ( max(sets == 3) )
      break
  }
  return ( tab )
}
```

### Wisconsin's Opponents

#### Wisconsin versus Colgate

```{r}
colgate = c(12, 15, 14)
wisconsin = c(25, 25, 25)

sum(wisconsin) / (sum(wisconsin) + sum(colgate))
```

- Wisconsin won 64.7% of all possible points

```{r}
colgate_theta = teams %>% 
  filter(Team == "Colgate") %>% 
  pull(theta)

wi_theta = teams %>% 
  filter(Team == "Wisconsin") %>% 
  pull(theta)

inv_logistic(wi_theta - colgate_theta)
```

- Wisconsin predicted to win 62.1% of all points on a neutral court.

```{r}
inv_logistic(wi_theta - colgate_theta + alpha)
```

- Wisconsin predicted to win 63.2% of points at home.

```{r}
set.seed(20211210)
sim_match(wi_theta - colgate_theta + alpha)
```


```{r}
N = 10000
result = rep(0,N)
wi_pts = rep(0,N)
opp_pts = rep(0,N)

for ( i in 1:N)
{
  m = sim_match(wi_theta - colgate_theta + alpha)
  result[i] = ( sum(m[1,] > m[2,],na.rm=TRUE) == 3 )
  wi_pts = sum(m[1,],na.rm=TRUE)
  opp_pts = sum(m[2,],na.rm=TRUE)
}

cat("Wisconsin versus Colgate\n")
cat(paste0("Wisconsin win % = ", mean(100*mean(result)),"\n"))
```

#### Wisconsin versus FGCU

#### Wisconsin versus UCLA


