---
title: "Volleyball and Modeling (continued)"
output:
  html_document: default
---

\newcommand{\given}{\,|\,}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning = FALSE,
                      cache=TRUE, autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(scales)
library(tidymodels)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

```{r default-theme, include=FALSE, eval=TRUE}
## reset default theme
theme_set(theme_bw())
```

### Volleyball Review

- There are 340 NCAA Division I Women's volleyball teams
- 64 teams made the NCAA tournament
- Wisconsin, the Big 10 champion, is one of these 64
- Wisconsin won their first four matches in the tournament versus Colgate, Florida Gulf Coast, UCLA, and Minnesota
- Wisconsin is in the Final Four and plays #1 seed and undefeated Louisville next
- The other semi-final has Nebraska, who knocked off #2 Texas, versus Pittsburgh.
- The [full bracket](https://www.ncaa.com/brackets/volleyball-women/d1/2021) is available on the NCAA web site.

#### Rules

- The winner of a volleyball match is the first to win three sets
- The winner of a set is the first to score the target or higher with a lead of at least two
- Sets 1--4 have a target of 25, a set 5 if needed has a target of 15.
- Each set is a series of rallies where one of the two teams wins a point.

#### Modeling Review

- We model each volleyball match as a series of binomial trials with the same probability that a given team wins the point versus the other team.
- Probabilities are based on a logistic regression model to predict the probability of winning a point based on the difference is strengths between the two teams.
- The parameters are:
    - $\theta_i$, the strength of team $i$ for teams $i = 1, \ldots, m$
    - A constraint is that $\sum_{i=1}^m \theta_i = 0$
    - $\alpha$, the added benefit for a home team
- The data is:
    - $x_k$ is the number of points won by the first team in match $k$
    - $y_k$ is the number of points won by the second team in match $k$
- The probability that team 1 wins match $k$ is
$$
p_k = \frac{1}{1 + \exp(\theta_{1[k]} - \theta_{2[k]} + \alpha_k)}
$$
- where:
    - $\theta_{1[k]}$ is the strength of team 1 in match $k$
    - $\theta_{2[k]}$ is the strength of team 2 in match $k$
    - $\alpha_k$ is the home field advantage is match $k$
        - $\alpha_k = \alpha$ if team 1 is the home team
        - $\alpha_k = -\alpha$ if team 2 is the home team
        - $\alpha_k = 0$ if neither team is the home team
- The log likelihood is the sum of log-likelihoods over all matches

$$
\ln L = \sum_{k=1}^n (x_k \ln p_k + y_k \ln (1-p_k))
$$

#### Inverse Logistic Function

- The key formula in the model is the *inverse logistic function* which takes an argument between $-\infty and \infty$ and returns a probability between $0$ and $1$.
    - The s-shaped function is increasing from a value of 0 at $-\infty$ and 1 at $\infty$.
    - The function rises from close to zero to close to one between $-5$ and $5$ as seen in this graph.
    
```{r logistic-curve, cho = FALSE, fig.height = 2}
inv_logistic = function(x) { return ( 1/(1 + exp(-x)) )}
delta = seq(-5,5,length.out=1001)
p = inv_logistic(delta)
df = tibble(delta,p)

ggplot(df, aes(x=delta,y=p)) +
  geom_line(color="blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 1, color="red", linetype = "dashed")
```  
    
    
### Data

- Data for all NCAA Division I volleyball matches is found on [the NCAA website](https://stats.ncaa.org/season_divisions/17720/scoreboards?utf8=%E2%9C%93&season_division_id=&game_date=12%2F09%2F2021&conference_id=0&tournament_id=&commit=Submit)
- There is a separate URL to produce a table with results for each date.
- I scraped the data and created a file with one row per match between Division I team
- The code block below reads in this data and eliminates matches from the national tournaments.

```{r}
vb_orig = read_csv("../../data/vb-division1-2021-all-matches-corrected.csv")

vb_all = vb_orig %>% 
  mutate(site = case_when(
    str_detect(site, "NCAA DI Volleyball Championship") ~ "NCAA DI",
    str_detect(site, "National Volleyball Invitational Championship") ~ "NCAA Invitational",
    TRUE ~ site))

vb = vb_all %>% 
  filter(is.na(site) |
           (!str_detect(site, "NCAA Invitational") &
               !str_detect(site, "NCAA DI")))

vb_ncaa = vb_all %>% 
  filter(site == "NCAA DI")

n_matches = nrow(vb)

teams_orig = read_csv("../../data/vb-division1-2021-teams.csv")

teams = teams_orig %>% 
  select(Team) %>% 
  mutate(team_index = row_number()) %>% 
  relocate(team_index)

conferences = teams_orig %>% 
  select(Team, Conference) %>% 
  mutate(team_index = row_number()) %>% 
  relocate(team_index)

n_teams = nrow(teams)

seeds = read_csv("../../data/vb-division1-2021-ncaa-seeds.csv")
```

- The (mostly) cleaned data set has `r n_matches` matches involving only these `r n_teams` division I teams from the regular season or conference tournaments.


### Data Set-up

- To wrangle the data in order to fit the model, we need to add index variables for each team.
- We also want to replace the scores over sets with the total number of points that each team scores.
- This block of code accomplishes this.
- We do this for the matches before the national tournament and those in the tournament

```{r data}
## all games
total_points_all = vb_all %>%
  mutate(x = s1_1 + s1_2 + s1_3 + 
           if_else(is.na(s1_4), 0, s1_4) +
           if_else(is.na(s1_5), 0, s1_5),
         y = s2_1 + s2_2 + s2_3 + 
           if_else(is.na(s2_4), 0, s2_4) +
           if_else(is.na(s2_5), 0, s2_5),
         home = as.integer(is.na(site))) %>%
  left_join(teams, by = c("team1" = "Team")) %>% 
  rename(index_1 = team_index) %>% 
  left_join(teams, by = c("team2" = "Team")) %>% 
  rename(index_2 = team_index) %>% 
  select(index, index_1, team1, conference1, index_2, team2, conference2, winner, loser, x, y, home, site)  

## all games before the national tournaments
total_points = total_points_all %>% 
  filter(is.na(site) |
           (!str_detect(site, "NCAA Invitational") &
               !str_detect(site, "NCAA DI")))

## games so far in the national tournament
total_points_ncaa = total_points_all %>% 
  filter(str_detect(site, "NCAA DI"))

## games including the national tournament
tp = bind_rows(total_points, total_points_ncaa)
```

### Maximum Likelihood

- This model includes a strength parameter $\theta_i$ for each team constrained so that $\sum_{i=1}^n \theta_i = 0$ and a single parameter $\alpha$ which is the home court advantage for a team if there is one team is the home team.
- This code uses the function `optim()` to maximize the parameters.

#### Code to fit the ML Model

```{r mle}
inv_logistic = function(x) { return ( 1/(1 + exp(-x)) ) }

calc_logl = function(theta, alpha, df)
{
  delta = theta[df$index_1] - theta[df$index_2] - alpha*df$home
  p = inv_logistic(delta)
  return ( sum( df$x*log(p) + df$y*log(1-p)) )
}

f = function(alpha_theta, df)
{
  alpha = alpha_theta[1]
  theta = c(alpha_theta[-1], -sum(alpha_theta[-1]))
  return ( calc_logl(theta, alpha, df) )
}

 
fit_ml_model = function(df)
{
  n_teams = df %>% 
    pull(index_1) %>% 
    max()
  theta_0 = rep(0, n_teams)
  out = optim(theta_0, f, method="BFGS",
            control = list(fnscale=-1),
            df = df)
  
  return ( out )
  
}
```

#### Fit the Model

```{r}
out = fit_ml_model(total_points)

alpha = out$par[1]
theta = c(out$par[-1], -sum(out$par[-1]))
teams = teams %>%
  mutate(theta = theta) %>% 
  left_join(conferences) %>% 
  relocate(Conference, .after = Team)
```

#### Output

```{r explore}
ranked_teams = teams %>%
  arrange(desc(theta)) %>%
  mutate(rank = row_number()) %>%
  left_join(seeds) %>% 
  relocate(rank, team_index) %>% 
  mutate(Conference = reorder(Conference, theta))
```

```{r}
ranked_teams %>% print(n=20)
```

#### Seed versus Rank

- The model's 15 strongest teams were all seeded, but the order of seeding differs from the model ranks.
- Creighton is the only seeded team the model does not include among the top 16 (they are ranked at #20).
- Wisconsin is ranked third, but is the #4 seed.
- The model and seeding agree on the top two teams, but in different order.
- #14 seed Creighton is ranked #20 by the model, but has an estimated $\theta = 0.37$ compared to $0.40$ for the models 16th best team, Penn State.

#### Theta Values of Seeded Teams

```{r}
ranked_teams %>% 
  filter(!is.na(Seed)) %>% 
ggplot(aes(x = Seed, y = theta)) +
  geom_col(color = "black", fill = "red")


ranked_teams %>% 
  filter(!is.na(Seed)) %>% 
  arrange(Seed) %>% 
  select(Team, Conference, theta, Seed, rank)
```

#### Wisconsin Tournament Opponents

```{r}
## Teams Wisconsin has beaten
ranked_teams %>%
  filter(Team %in% c("Wisconsin", "Colgate", "FGCU", "UCLA", "Minnesota")) %>% 
  arrange(desc(theta))

## Teams Wisconsin Might Play in the Final Four
ranked_teams %>%
  filter(Team %in% c("Wisconsin", "Louisville", "Nebraska", "Pittsburgh")) %>% 
  arrange(desc(theta))
```

## Compare Results with the Data

- For the 60 NCAA matches which have occurred, compare the actual proportions of points won with what the model predicts

```{r}
ncaa = total_points_ncaa %>% 
  select(team1, team2, winner, loser, index_1, index_2, x, y) %>% 
  mutate(theta_1 = theta[index_1],
         theta_2 = theta[index_2],
         p_model = inv_logistic(theta_1 - theta_2),
         p_observed = x/(x+y)) %>% 
  select(-index_1, -index_2)
```

```{r, include = FALSE}
caption_string = "**NCAA Results**. Each points represents one of the matches of the 2021 NCAA Division I Volleyball Tournament. The value on the x axis is the probability that one team wins a point as predicted by the model. The value on the y axis is the observed proportion of points won by this team. The blue line is a fitted regression line. The red line is y = x."
```

```{r, fig.cap = caption_string}
ggplot(ncaa, aes(x = p_model, y = p_observed)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  geom_point() +
  xlab("Modelled Team 1 Point Probability") +
  ylab("Observed Team 1 Point Proportion") +
  ggtitle("NCAA Tournament Model versus Results")
```

- There is reasonable agreement between the model's predictions about the probability that the first team will win a point and the observed proportion.
- The first team listed is the underdog by the tournament seeding.

### Simulation

- We next explore simulation to calculate the probability of winning a match
- For any two teams, the model produces an estimate of the probability that each team wins a point
- By simulation, we determine the probability that each team wins the match.

```{r simulation, include = FALSE}
inv_logistic = function(x) { return ( 1/(1 + exp(-x)) )}

## simulate a single set to the target score of 25 or 15
sim_set = function(delta,target)
{
  p = inv_logistic(delta)
  points = c(0,0)
  repeat
  {
    pt = rbinom(1,1,p)
    if ( pt == 1 )
      points[1] = points[1] + 1
    else
      points[2] = points[2] + 1
    if ( max(points) >= target && abs(diff(points)) >= 2 )
      break
  }
  return ( points )
}

## simulate a match by simulating the sets
sim_match = function(delta)
{
  tab = matrix(NA,2,5)
  sets = c(0,0)
  index = 1
  repeat
  {
    if ( sum(sets) < 4 )
      result = sim_set(delta,25)
    else
      result = sim_set(delta,15)
    if ( result[1] > result[2] )
      sets[1] = sets[1] + 1
    else
      sets[2] = sets[2] + 1
    tab[,index] = result
    index = index + 1
    if ( max(sets == 3) )
      break
  }
  return ( tab )
}

sim_matches = function(delta=0.5, N = 1000)
{
  result = rep(0,N)
  team1_pts = rep(0,N)
  team2_pts = rep(0,N)

  for ( i in 1:N)
  {
    m = sim_match(delta)
    result[i] = ( sum(m[1,] > m[2,],na.rm=TRUE) == 3 )
    team1_pts = sum(m[1,],na.rm=TRUE)
    team2_pts = sum(m[2,],na.rm=TRUE)
  }
  
  return ( mean(result) )
}
```

### 1000 Simulates of each NCAA match

```{r}
set.seed(20211213)
delta = (ncaa %>% pull(theta_1)) - 
  (ncaa %>% pull(theta_2))

ncaa_match_prob = map_dbl(delta, sim_matches)
```

### Store results 

```{r}
ncaa = ncaa %>% 
  mutate(model_prob = ncaa_match_prob,
         winner_prob = case_when(
           team1 == winner ~ model_prob,
           TRUE ~ 1 - model_prob))

## Plot the results
ggplot(ncaa, aes(x = winner_prob)) +
  geom_histogram(boundary = 0.5, binwidth = 0.05,
                 color = "black", fill = "skyblue") +
  xlab("Model Probability of Winning the Match") +
  ylab("# of Matches") +
  ggtitle("2021 NCAA DI Volleyball")
```

```{r}
ncaa %>% 
  select(winner, loser, winner_prob) %>% 
  arrange(winner_prob) %>% 
  print(n = Inf)
```

- There were 10 upsets from the first 60 matches, by the model

### Results of Matches from Seeded Teams

```{r}
ncaa = ncaa %>% 
  left_join(seeds, by = c("winner" = "Team")) %>% 
  rename(winner_seed = Seed) %>% 
  left_join(seeds, by = c("loser" = "Team")) %>% 
  rename(loser_seed = Seed)

ncaa_seeded = ncaa %>% 
  filter(!is.na(winner_seed) | !is.na(loser_seed)) %>% 
  select(winner_seed, winner, loser_seed, loser, winner_prob)

ncaa_seeded %>% 
  print(n = Inf)

ncaa_seeded %>% 
  filter(!is.na(loser_seed))
```

- In the second round matches involving seeded teams, the only upset was Creighton losing to Kansas.
- The model had this as a large upset, giving Kansas only a 15.5% chance of winning, but Creighton is the only seeded team that the model said was undeserving to be seeded.
- Penn State, which the model would have seeded, lost to #3 seed Pittsburgh in the second round.

### Predictions for the Final Four

#### Wisconsin versus Louisville

```{r}
wi_theta = ranked_teams %>% 
  filter(Team == "Wisconsin") %>% 
  pull(theta)

wi_theta

louisville_theta = ranked_teams %>% 
  filter(Team == "Louisville") %>% 
  pull(theta)

louisville_theta

delta = wi_theta - louisville_theta 

inv_logistic(delta)

sim_match(delta)

sim_matches(delta, N = 10000)
```


#### Pittsburgh versus Nebraska

```{r}
pitt_theta = ranked_teams %>% 
  filter(Team == "Pittsburgh") %>% 
  pull(theta)

pitt_theta

neb_theta = ranked_teams %>% 
  filter(Team == "Nebraska") %>% 
  pull(theta)

neb_theta

delta = pitt_theta - neb_theta 

inv_logistic(delta)

sim_match(delta)

sim_matches(delta, N = 10000)
```
