---
title: "Background"
author: "Kai Tsuyoshi"
date: "11/27/2021"
output: html_document
---
Introduce the background and one or more general questions of interest
Answer: average length of hurricanes (time dissipated - time formed) by season (setting the season by date formed rather than dissipated)

Describe data with information about these questions
Describe the data source, who collects it, how it is made available, and so on.
Answer: The data source we’re using was created by using a python program to extract data from a Wikipedia page about hurricanes (https://en.wikipedia.org/wiki/List_of_United_States_hurricanes). The data was collected for a project from CS220 about hurricanes. We got this from https://github.com/msyamkumar/cs220-f21-projects/tree/main/p6 and it was made available by the instructors.

Provide a description of key variables in the data you will examine to address these questions
This could be in a paragraph or a table
	Answer: First, we’ll mutate to create two columns, one for length between formed and dissipated and one for season. Using this information, we can find the average length of hurricanes by season
 Describe a connection between the data and one or more refined questions of interest





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 3)
library(tidyverse)
library(lubridate)
library(scales)
source("../../scripts/viridis.R")
```


```{r}
hurricanes = read_csv("../../data/hurricanes.csv")

hurricanes = hurricanes %>% 
  mutate(formed = as.Date(formed, "%m/%d/%Y")) %>% 
  mutate(dissipated = as.Date(dissipated, "%m/%d/%Y")) %>% 
  mutate(year = year(formed), month = month(formed, label = TRUE), day = mday(formed), wday = wday(formed, label = TRUE)) %>% 
  mutate(season = case_when(month %in% c("Mar", "Apr", "May") ~ "Spring",
            month %in% c("Jun", "Jul", "Aug") ~ "Summer",
            month %in% c("Sep", "Oct", "Nov") ~ "Fall",
            month %in% c("Dec", "Jan", "Feb") ~ "Winter"))

hurricanes
```


Here is a graph 
```{r}
hurricanes$length = difftime(hurricanes$dissipated, hurricanes$formed, units = c("days"))



hurricanes = hurricanes %>% 
  mutate(length = as.numeric(length)) %>% 
  arrange(month, season) 
  
seasonal = hurricanes %>% 
  mutate(month = match(month,month.abb))


ggplot(seasonal, aes(x = month, y = length, fill = season)) +
  geom_col() +
  facet_wrap(~season)
 


```


