---
title: "project_prop"
author: "Kai Tsuyoshi"
date: "11/20/2021"
output: html_document
---
## Project Proposal


## Group 311C: Kai Tsuyoshi, Fei Qin, William Fan, Jason Fan, Nihar Simhadri


## Communication Plan: 
- Use email for document and dataset attachments back-and forth
- For general meetings to collaborate, Zoom will be used if the group cannot meet in person
- Main meetings during or after discussion sections every Monday
- Get-stuff-done meetings on Fridays (for meeting deadlines, discussing issues, etc.)



## Questions of Interest
- What season (split dates up into four seasons) tends to have the most occurrences of hurricanes?
- average length of hurricanes (time dissipated - time formed) by season
- mean speed?
- most damage / death?
- hypothesis test for likelihood of a repeat of the most destructive hurricane?

## Data Source
- CSV was created by extracting data from https://en.wikipedia.org/wiki/List_of_United_States_hurricanes

## Data Description
- Is a dataset of hurricanes that made landfall in the US
- Variables are as follows: name(chr), date formed(chr), date dissipated(chr), mph(dbl), damage(chr) (monetary value), deaths(dbl)
- 141 observations
- Each observation is a hurricane that made landfall with the above variables

## Graph 
- facet_wrap() graph showing occurrences of hurricanes by season or by length
  - could be color-coded based on name or another variable


## Methods
- A regression model of damage and/or deaths 
- Compare distribution to a normal curve??

```{r}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 3)
library(tidyverse)
library(lubridate)
library(scales)
source("../../scripts/viridis.R")
```


```{r}
hurricanes = read_csv("../../data/hurricanes.csv")

hurricanes

```

