---
title: "Chimpanzees Analysis and Hypothesis Tests"
author: "Bret Larget"
output: html_document
---

This R Markdown document includes contributions from Jessi Kehe.

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}

### Setup details

- This lecture use the following scripts, assumed to be in your course scripts directory.
    - `COURSE/scripts/viridis.R`
    - `COURSE/scripts/ggprob.R`
    
- You also need the chimpanzee data
    - `COURSE/data/chimpanzee.csv`

- You will need the package `tidyverse` for these lectures.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      cache=TRUE,
                      autodep=TRUE,
                      cache.comments=FALSE)
library(tidyverse)
library(lubridate)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```


## Overview of Hypothesis Tests for p

## Recall:  Case Study

- Chimpanzees were studied in a pro-social choice experiment.
    - Chimpanzees are given a sequence of choices about which color token to select.
    - One color (labeled selfish) results in them getting food.
    - Another (labeled pro-social) results in both them and a partner chimp receiving food.
    - In a control setting, there is no partner
        - different color choices result in the researcher mimicking the same actions as when partner is present (i.e., always giving the actor chimp food, making a motion with food toward the empty room for the partner chimp).
- We will develop a statistical model for this experiment.
- Each part of the experiment includes a series of 30 trials where one of two outcomes is possible.


## Recall:  Inference questions

- How often does the actor chimpanzee make the pro-social choice?
- How much uncertainty is there in this estimate?
- How confident can we be that the long-run probability of making the pro-social choice in repeated trials falls within some interval?
- Do chimps make the pro-social choice more than 50 percent of the time, or is the pro-social choice made at random?
- Does the frequency with which the chimpanzee makes the pro-social choice depend on there being a partner chimpanzee in the neighboring room?


## Recall:  Partial Data

- One chimpanzee (subject A) was tested 90 times with a partner (three sets of 30 trials with each of three different partners)
- The chimpanzee made the pro-social choice 60 times out of 90 trials.

## Recall:  A Statistical Model

1. Name and define variables
2. Make assumptions (these can be examined later)
3. Write a probability model for the variables and required parameters

- $X$ is the number of pro-social choices that subject A makes in 90 trials.
- Assume:
    - each trial could be pro-social or selfish
    - 90 trials was predetermined
    - trials are independent (this assumption is debatable)
    - same probability of pro-social choice for each trial (this assumption is debatable)
- Under these assumptions, $X$ has a binomial distribution.
- Parameters are:
    - $n$, the number of trials. Here $n=90$.
    - $p$, an unknown long-run probability that this chimpanzee makes the pro-social choice. (We may be interesting in inference on this parameter using the data.)
    
The model is:

$$
X \mid p \sim \text{Binomial}(n,p)
$$

The data are:

- $n=90$
- $x=60$

## Hypothesis Tests

- When finding a confidence interval, we are trying to estimate an unknown parameter, such as $p$.
- For a *hypothesis test*, we often examine the data to see if the data are *consistent with  the parameter having a specific fixed value* versus an alternative where it is different.
- Typically, the null hypothesis represents a condition of no effect.
- In the chimpanzee example, it makes sense to pose these null hypothesis and alternative hypotheses.
    - $H_0: p = 0.5$
    - $H_a: p \neq 0.5$
      
- The null hypothesis is what we would expect if the chimpanzee were choosing colored tokens at random.
- The alternative hypothesis is the long-run probability is something other than 0.5.
    - This difference may be interpreted as willful behavior by the chimpanzee to do something (whether acting in a pro-social or selfish manner more often than expected by chance).


## Hypothesis Tests for p

### Hypothesis Test Steps

1. State the statistical model for the data
2. State hypotheses
3. Choose a test statistic
4. Determine the sampling distribution of the test statistic *when the null hypothesis is true*.
5. Determine which outcomes are *at least as extreme as the observed test statistic*, or *which outcomes are at least as favorable to the alternative hypothesis as the observed test statistic* and find the collective probability of these outcomes. This probability is called a *p-value*.
6. Use the p-value to interpret the strength of evidence against the null hypothesis.
    - Conventional choices are to call:
        - $p < 0.05$ *statistically significant*;
        - $p < 0.01$ *highly statistically significant*.
7. Interpret the result in context, summarizing the statistical evidence by referring to the p-value and test.

**Note: use context to determine if $p$ is the parameter of the statistical model (the long-run proportion of time the chimpanzee makes the pro-social choice) versus a p-value (the probability of obtaining the observed data or something more extreme in favor of the alternative hypothesis when assuming that the null hypothesis is true).**


## Example

Our chimpanzee made the pro-social choice 60 times out of 90.

1. Model:

$$
X \mid p \sim \text{Binomial}(90,p)
$$

2. State hypotheses:

$$
H_0: p = 0.5 \\
H_a: p \neq 0.5
$$

3. Test statistic is $X$. (You might have learned to use $\hat{p}$ in an AP Stat course.)

4. If the null hypothesis is true, then

$$
X \sim \text{Binomial}(90,0.5)
$$

5. Any outcome as likely or less likely than observing $X=60$ would provide evidence against $p=0.5$ at least as strongly as the observed value.

- The $\text{Binomial}(90,0.5)$ distribution has mean 45 and is symmetric, which suggests outcomes 30 or smaller or 60 or higher have the same or smaller probability as $X=60$ under the null hypothesis.

- This is the distribution we assume is true (based on the null hypothesis).

```{r graph-ho, fig.height=3}
gbinom(90, 0.5, scale=TRUE) +
  geom_vline(xintercept=60, color="red",
             linetype="dashed") +
  
  theme_bw()
```

- Here are close-up looks in the left and right tails.

```{r, fig.height = 3}
## left tail
gbinom(90, 0.5, a=0, b=30) +
  geom_binom_density(90, 0.5, a=31, b =32, color = "gray") +
  geom_hline(yintercept = dbinom(60,90,0.5), color = "red", linetype = "dashed") +
  theme_bw()

## righttail
gbinom(90, 0.5, a=60, b=90) +
  geom_binom_density(90, 0.5, a=58, b =59, color = "gray") +
  geom_hline(yintercept = dbinom(60,90,0.5), color = "red", linetype = "dashed") +
  theme_bw()
```

- Next, we calculate the p-value.
    - Verify first that the individual outcomes
$X = 0, 1, \ldots, 30, 60, \ldots, 90$ and $X \ge 60$ are those that have the same or lower probability of occurring if $p = 0.50$

- To avoid checking equality for continuous values (rounding errors could lead to two values that are analytically equal being regarded as unequal), we introduce a small tolerance and look for $x$ values whose probability is less than the probability of exactly 60 successes plus this small tolerance.
- I use the **dplyr** function `near()` to do so.

```{r extreme-x}
## tidyverse calculation to find all x
##    such that dbinom(x, 90, 0.5) <= dbinom(60, 90, 0.5)

p_60 = dbinom(60, 90, 0.5)

temp = tibble(
  x = 0:90,
  prob = dbinom(x, 90, 0.5)) %>% 
  filter(prob < p_60 | near(prob, p_60)) 

## pvalue sums the probabilities
##   for x = 0, 1, ..., 30, 60, 61, ..., 90

pvalue = temp %>% 
  summarize(pvalue = sum(prob)) %>% 
  pull(pvalue)

pvalue

## more direct calculation assuming that c(0:30,60:90) is correct
pvalue2 = sum(dbinom(c(0:30,60:90), 90, 0.5))
pvalue2

## yet another calculation using pbinom
## recall P(X >= 60) = 1 - P(X <= 59)
pvalue3 = pbinom(30, 90, 0.5) + (1 - pbinom(59, 90, 0.5))
pvalue3
```

6. The p-value is about `r signif(pvalue,2)` which is about 1 in `r round(1/pvalue)`. This is highly statistically significant.

- If the null hypothesis were true,
we would expect to see a result this extreme only about once per every 485 experiments, yet it happened.
- There is strong evidence that the null hypothesis is false and that the true value of $p$ is closer to the observed proportion of $\hat{p} \approx 0.667$.  
     - It is more plausible that $p$ is not 0.5 and we just observed typical data, than it is that $p=0.5$ and we just witnessed a very improbable outcome.

7. Draw a conclusion *in context, without using technical jargon*. Summarize the statistical evidence to support the evidence in parentheses.

> There is very strong evidence ($p \doteq 0.0021$, two-sided binomial test) that the chimpanzee in this experiment will make the pro-social choice more than half the time in the long run under similar experimental conditions.

- Note that the conclusion:
    - speaks about the strength of evidence and not certainty of the conclusion
    - explains what the conclusion means in the context of the experimental setting
    - has a brief statistical summary of the evidence that:
        - is in parentheses
        - reports the calculated p-value without too much precision
        - states if the test is one- or two-sided, if appropriate
        - names the test used
- In this case, the test is calculated by computing an exact binomial probability, and hence is named a *binomial test*.  

- Note that the conclusion in context **does not**:
    - Say that the null hypothesis is accepted or rejected
    - Use technical terms such as null or alternative hypothesis

- We are **not making a decision about truth**.
- We are stating **a measure of the strength of the statistical evidence** against a particular hypothesis in favor of a specific alternative
- By reporting the p-value, we let the reader use their own measure of how improbable an outcome would need to be in order to be convincing
- The statistical hypothesis test depends on all of the modeling assumptions that we made.
- A decision should be based on:
    - the consequences of the decision;
    - the statistical evidence;
    - an assessment of the effects if the assumptions are not fully met
    
**When asked to interpret the results of a hypothesis test in context, do not respond:**

> We reject the null hypothesis.

> The null hypothesis is not true.

> We accept the null hypothesis.

> The test is significant.

- If a question asks, what decision would be made in a hypothesis test for a
significance level, say, $\alpha = 0.05$, then if the p-value is less than this significance level, examples of appropriate responses would be:

> The hypothesis test is rejected at the 0.05 level of significance

> The hypothesis test is not rejected at $\alpha = 0.05$









## More on hypothesis testing 

- To carryout a hypothesis test, a test statistic is specified which allows us to measure the compatibility between the null hypothesis and the data.  
     - In our previous chimpanzee example, the chimpanzee made the pro-social choice 60 times out of 90.  Our test statistic was $x = 60$ (the observed realization of our binomial random variable).  We were then able to measure the compatibility between $X=60$ and the null hypothesis with $p = .5$.
     
- The p-value contributes to quantifying this compatibility.  
    - The p-value is the probability of observing what we did for the test statistic (e.g., $X=60$) or something more extreme in the direction of the alternative hypothesis.
    - In the previous chimpanzee example, we considered the *two-sided* alternative of $p \neq 0.5$ so we looked to both tails to estimate the p-value.  
        - If we used $H_a: p > 0.5$, then we would be looking in the *upper* tail to compute the p-value ($\prob(X \geq 60)$)  
        - If we used $H_a: p < 0.5$, then we would be looking in the *lower* tail to compute the p-value ($P(X \leq 60)$); note that this would result in a p-value > 0.50.  
    - Smaller p-values are evidence *against* the null hypothesis.  

- Failing to find statistical significance means the null hypothesis is *not* rejected.  Note that this is not the same as *accepting* the null hypothesis.

#### Simple example

> A noted psychic was tested for extrasensory perception (ESP). The psychic was presented with 200 cards face down and asked to determine if the cards were one of five symbols: a star, a cross, a circle, a square, or three wavy lines. The psychic was correct in 50 cases. Let p represent the probability that the psychic correctly identifies the symbol on the card in a random trial. Assume the 200 trials can be treated as a simple random sample from the population of all guesses the psychic would make in his lifetime.  Specify a hypothesis test to determine if this psychic did better than random guessing, and carryout the hypothesis test with the sample. [These numbers are made-up and are not based on an actual study.]


The psychic made the correct choice 50 times out of 200.

1. Model: $X \mid p \sim \text{Binomial}(n,p)$

2. State hypotheses:

$$
H_0: p = 0.2 \\
H_a: p > 0.2
$$

3. Test statistic is $X$. 

4. If the null hypothesis is true, then

$$
X \sim \text{Binomial}(200,0.2)
$$

5. Outcomes great than or equal to $X=50$ would provide evidence against $p=0.2$.

The $\text{Binomial}(200,0.2)$ distribution has mean 40.

This is the distribution we assume is true (based on the null hypothesis).

```{r graph-esp, fig.height=3}
gbinom(200, 0.2, scale = TRUE) +
  geom_vline(xintercept = 50, color = "red",
             linetype = "dashed", alpha = 0.5) +
  theme_minimal()
```


- Next, we calculate the p-value.

```{r esp-pvalue}
## P(X >= 50)
p_value = 1 - pbinom(49, 200, 0.2) # P(X > 49) = P(X >= 50)
p_value
```

- The p-value is `r round(p_value,4)`, which is borderline significant.  If we set the significance level to 0.05, we would reject the null hypothesis that the psychic was randomly guessing.

Here is a conclusion in context:

> There is moderate evidence that the psychic can select the correct symbol more frequently than 20%, the rate when guessing at random ($p = 0.049$, one-sided binomial test).




