---
title: "STAT 240 Discussion 11 Solution"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

## Assignment Completion

If your group does not complete the assignment during discussion,
find time to work together outside of class to finish the assignment.

Group 311C: William Fan, Jason Fu, Fei Qin, Nihar Simhadri, Kai Tsuyoshi

## Data

The data set `boston-marathon-2010-sample.csv` contains a random sample of 50 men and 50 women, aged 35--39, who ran in and completed the 2010 Boston Marathon.
Times are in minutes.

```{r}
bm = read_csv("../../data/boston-marathon-2010-sample.csv")
```

## Problems

The questions in this discussion assignment use a subset of the 2010 Boston Marathon data set described above.

### Problem 1

Display the data using side-by-side boxplots with the sex of the competitor on the x axis and their time on the y axis.
Add an appropriate title and axis labels to the plot.

```{r}
ggplot(bm, aes(x = Sex, y = Time)) +
  geom_boxplot() +
  ggtitle("Distribution of Marathon Times based on Sex") +
  xlab("Sex") +
  ylab("Time (in Minutes")
```

### Problem 2

Create and display a table with a numerical summary of the table.
Calculate the sample size, mean time, standard deviation of the times, and standard error of the mean ($s_x/\sqrt{n}$) separately for each sex.
This table will have one row for each sex.

```{r}
tbl = bm %>% 
  group_by(Sex) %>% 
  summarize(n = n(), mean = mean(Time), sd = sd(Time), se = sd/sqrt(n))

tbl
```


### Problem 3

Using the function `t.test()`, construct separate 95% confidence intervals for each sex for the mean time to complete the 2010 Boston Marathon among all competitors of that sex aged 35--39.

```{r}
# male
x = bm %>% 
  filter(Sex == "male") %>% 
  select(Time)


t.test(x)
```



```{r}
# female
  y = bm %>% 
  filter(Sex == "female") %>% 
  select(Time)


t.test(y)
```

### Problem 4

Without using the function `t.test()` write code to calculate the end points of the 95% confidence interval for the mean time for women aged 35--39 to complete the 2010 Boston Marathon.
$$
t = \frac{\bar{X} - \bar{Y}}{\sqrt{\frac{s_x^2}{n_x} + \frac{s_y^2}{n_y}}}
$$

```{r}
t_mult = qt(0.975, 50-1)

female_mean = tbl %>%
  filter(Sex == "female") %>% 
  select(mean)

female_se = tbl %>%
  filter(Sex == "female") %>% 
  select(se)

ci = c(female_mean - t_mult * female_se, female_mean + t_mult * female_se)
ci



```

### Problem 5

Write an interpretation of the confidence interval from the previous problem in context, following course examples.

For the sample size of n = 50, we are 95% confident that the mean time for women aged 35--39 to complete the 2010 Boston Marathon is between 230.7 minutes and 253.2 minutes.

### Problem 6

Test the hypothesis that the mean times are equal for men and women to complete the 2010 Boston Marathon among runners aged 35--39 versus the two-sided alternative that they are not.

1. Define parameters and a model
2. State hypotheses
3. Use `t.test()` to do calculations of the test statistic and p-value.

```{r}
female_mean = bm %>%
  filter(Sex == "female") %>% 
  select(Time)
male_mean = bm %>%
  filter(Sex == "male") %>% 
  select(Time)
difffe=female_mean-male_mean
meand=mean(difffe$Time)
result=t.test(difffe, mu=meand)
result
```

### Problem 7

Interpret the results of this hypothesis test in context, following examples from lecture.

There is not enough evidence that disproves the hypothesis.

### Problem 8

Create a graph using functions from `ggprob.R` to graph the t distribution which is the null distribution of the test statistic.
Shade the area under the curve that represents the p-value and add a vertical dashed line at the location of the test statistic.

```{r}
gt(df = 49, a = 17, b = 40) +
  geom_t_fill(df = 49, a = 28.584, b = 40) +
  geom_vline(xintercept = 28.584, color = "red", linetype = "dashed") +
  xlab("t") +
  theme_minimal()
```







