---
title: "Getting Started with stringr"
author: "Bret Larget"
output: html_document
##editor_options: 
##  chunk_output_type: outline
---

This R Markdown document includes contributions from Professor Jessi Kehe. 

### Setup details

- You will need the packages **tidyverse** including **stringr** within.

- Be sure to read *R4DS* Chapter 14 for information on **stringr** 

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, autodep = TRUE,
                      cache.comments = FALSE)
library(tidyverse)
```


## Strings and Regular Expressions

## Overview

- This lecture will give examples of working with strings and regular expressions.


## Strings

- A *string* is a sequence of characters.
- In R, a string is surrounded by a pair of single quotes or double quotes.

```{r}
'a string'
"another string"
```

### Quotes in strings

- A string which contains quotes may be surrounded by the other type.

```{r}
"A 'string' that contains single quotes"
'A "string" that contains double quotes'
```

- Alternatively, use a backslash (`\`) before a quote character, or any other special character.

```{r}
"A \"string\" using backslashes to \"escape\" the double quote characters."
```

### Common string functions

- See the Cheat Sheet for a more extensive list.
- All of the string functions begin `str_`.

### Detect matches

- A pattern is a string which holds a *regular expression* that can match strings with particular patterns
- The simplest pattern is an exact match with the string
- Later, we describe many ways to build regular expressions

#### Example functions

- `str_detect(string, pattern)` --- detect a pattern in a string
- `str_count(string, pattern)` --- count the number of times a string contains a pattern

### Subsets

- A subset is part of the string
- When selecting by position, positive numbers count from the beginning, negative count from th end.

#### Example functions

- `str_sub(string, start, end)` --- extract a subset by position
- `str_extract(string, pattern)` --- extract first match of pattern, `NA` if no match

### Length

- The *length* of a string is the number of characters in it.

#### Example functions

- `str_length(string)` --- length of the string
- `str_pad(string, width, side, pad = " ")` --- pad strings to a constant width

### Mutate

- Modify strings

#### Example functions

- `str_sub(string, start, end) = replacement_string` --- replace a subset by position
- `str_replace(string, pattern, replacement)` --- replace first match with replacement
- `str_to_lower(string)` --- change all letters to lower case
- `str_to_upper(string)` --- change all letters to upper case

### Join and split

- Combine strings or divide them

#### Example functions

- `str_c(..., sep = "")` --- combine multiple strings into one
- `str_split_n(string, pattern, n)` --- return the nth substring that matches the pattern

## Regular Expressions

- A *regular expression* is a sequence of characters that defines a pattern which may match a string or part of a string
- Most characters match only their matching character:
    - `a` matches only `a`, `A` matches only `A`
    - `,` and `;` match the corresponding punctuation marks
    - Usually, a digit such as `1` or `2` matches only the corresponding digit
        - An exception is when preceded by a backslash, `\1`
- There are many special characters that take on special meaning in a regular expression
- When you want the literal special character, you need to precede it with a backslash, `\`.

### Examples

#### Alternates

- `|` --- *or*, match exactly one of the possibilities: can separate more than two chocies
- `[]` --- *one of*, match one of the characters between the brackets
- `[^]`--- *anything but*, match anything *not* between `[^` and `]`
- ranges --- examples: `[a-d]`, `[0-5]`, `[a-zA-Z]`

#### Anchors

- `^` --- the start of the string
- `$` --- the end of the string

#### Quantifiers

- How often does the previous character/group need to appear?

- `?` --- 0 or 1
- `*` --- 0 or more
- `+` --- 1 or more
- `{n}` --- exactly `n` times
- `{n,}` --- `n` or more
- `{m,n}` --- form `m` to `n` times

### Special collections

- `[:space:]` --- space, tab, newline
- `[:blank:]` --- space and tab
- `[:punct:]` --- a collection of punctuation characters
- `[:symbol:]` --- a collection of symbols
- `[:digit:]` --- a digit
- `[:lower:]` --- same as `[a-z]`, a lower-case letter
- `[:upper:]` --- same as `[A-Z]`, an upper-case letter
- `[:alpha:]` --- same as `[a-zA-Z]`, `[:lower:]|[:upper]`
- `[:alnum:]` --- letter or number, `[:alpha:]|[:digit:]`
- `[:graph:]` --- punctuation, symbols, digits, letters
    - same as `[:punct:]|[:symbol:]|[:alnum:]`
- `.` --- matches any single character except for a newline    
    
#### Others

- See the cheat sheet for others:
    - Peek ahead and behind
    - Groups
    
## Regular expression representations in R

- Each regular expression is represented as a string in R
- Most regular expressions are simply the string of the regular with no changes
    - the regular expression `abc+` is represented as the string `"abc+"`
- However, regular expressions that include a backslash, `\`, need to duplicate the `\` in the string representation as `\` means something special inside a string.
    - "\\n" is the new line character
    - "\\t" is the tab character
    - and there are more examples

#### Examples

- The regular expression `a+`, one or more consecutive `a`s has the string representation `"a+"`
- The regular expression `a\+` for an `a` followed by a literal `+` has the string representation `"a\\+"`
- The regular expression `\\` to match a literal `\` has the string representation `"\\\\"` as each `\` in the regular expression requires `\\` in the string representation.

## Data

- We create a simple data frame consisting of the built in names in R.
    - `colors()` creates a vector of color names
    - `seq_along()` creates a sequence from 1 to the length of the argument.

```{r color-data}
df_colors = tibble(
  index = seq_along(colors()),
  colors = colors())

print(df_colors)
```

## Problems

> Find the shortest and longest color names

- It would be nice to have a function which returns the length of a string.
- Consult the **stringr** cheat sheet:
    - Aha! Use `str_length()` to find the length of each string  
- Then use `slice_max()` or `slice_min()` to identify the extremes  

```{r}

```





#### Solution

```{r}
## Add the string length
df1 = df_colors %>%
  mutate(length = str_length(colors))

## Longest
longest = df1 %>%
  slice_max(n = 1, order_by = length)

longest

## Shortest
shortest = df1 %>%
  slice_min(n = 1, order_by = length)

shortest
```

- The longest color name is `r longest$colors` with a length of `r longest$length` characters.  
- The shortest color names are `r shortest$colors[1]` and `r shortest$colors[2]`, each with a length of `r shortest$length[1]`.

> Make a bar plot of the string lengths

- Note use of `scale_x_continuous()` to change the default breaks and labels in the plot.

```{r length-plot}
ggplot(df1, aes(x=length)) +
  geom_bar(fill = longest$colors, color="black") +
  xlab("Word Length") +
  scale_x_continuous(labels = 3:20, breaks = 3:20) +
  ggtitle("R Colors Name Length Distribution") +
  theme_bw()
```


> Find the color names of length 8. Report the number, and the first and last color name (alphabetically).

```{r}

```







#### Solution

```{r length-8}
df2 = df_colors %>%
  filter(str_length(colors) == 8) %>% 
  arrange(colors)

df2 %>% 
  summarize(n = n(),
            first = min(colors),
            last = max(colors))

df2 %>% 
  print(n = Inf)
```

> How many color names begin and end with the same letter? Report the number and the first and last alphabetically.

```{r}

```





#### Solution 1

- One approach is to create new variables with the first and last letters of each column and then filter for cases where these are the same.
- Is there a function to grab part of a string? Check the cheat sheet:
- Aha! There are several potentially useful functions.
- `str_sub(string, start, end)` returns the sub string from `start` to `end`
    - Use negatives to count from the end.
    - Can also be used to assign values by position
- `str_extract(string, pattern)` returns the first substring to match the regular expression encoded in `pattern`.
    
```{r}
## Solution using str_sub()
## Positive start and end count from the start
## Negative count from the end
df3a = df_colors %>%
  mutate(first = str_sub(colors,1,1),
         last = str_sub(colors,-1,-1)) %>%
  filter(first == last)
df3a

df3a %>%
  summarize(n=n(), first = min(colors), last = max(colors))
```

#### Solution 2

- Next, use regular expressions and `str_extract()`.
- Note the special characters:
    - `^` matches the start of the string
    - `$` matches the end of the string
    - `.` matches any single character (except for end of line)

```{r}
## Solution using str_extract()
df3b = df_colors %>% 
  mutate(first = str_extract(colors, "^."),
         last = str_extract(colors, ".$")) %>% 
  filter(first == last)

df3b

df3b %>%
  summarize(n=n(), first = min(colors), last = max(colors))
```

#### Solution 3

- We could have accomplished the same thing without creating the extra columns
- However, it can be helpful to create the columns to make sure the code is doing what you think it is doing

```{r}
df_colors %>% 
  filter( str_extract(colors, "^.") == str_extract(colors, ".$") )
```

#### Solution 4

- Another fancy solution uses the group operator, `()`.
    - Each part of a regular expression surrounded by parentheses is saved and labeled for further use.
        - The first part may be recalled with `\1`, the second with `\2`, and so on.
        - When representing `\` in a regular expression as a string in R, recall it needs to be "escaped" with a leading `\`
- We use `str_detect()` to detect when the color name matches a pattern where the first character is a letter that is also used as the last character.
- Note the use of the special regular expression syntax `[:alpha:]` which matches letters of the alphabet.
    - Alternatives are:
        - `[:lower:]` or `[a-z]` for lowercase letters
        - `[:upper:]` or `[A-Z]` for uppercase letters
        - `[a-zA-Z]` for lowercase or uppercase letters
- The character `.` matches any character.
- The character `*` repeats the previous pattern 0 or more times.
- Put all together, the regular expression `^([:alpha:]).*\1$` matches:
    - a letter at the beginning of the string
    - then any set of characters of length 0 or more
    - followed by the same letter as at the start of the string at the end
- We need to replace `\` with `\\` in the string representation of the regular expression.

```{r}
df3c = df_colors %>%
  filter(str_detect(colors, "^([:alpha:]).*\\1$"))

df3c
```


> Find all color names that begin with a letter from "aeiou" and end with a letter from "aeiouy". Modify so that the first letter is capitalized and the last two characters are each replaced by X.

```{r}
prob4 = df_colors %>% 
  filter(str_detect(colors, "^[aeiou].*[aeiouy]$")) 

prob4 = prob4 %>% 
  mutate(colors = str_to_title(colors)) %>% 
  mutate(colors = str_c(str_sub(colors, 1, -3), "XX")) 
prob4
```






#### Solution

- Regular expression special symbols:
    - `^` is the beginning of the string
    - `$` is the end of the string
    - `.` is any character (except a newline)
    - `*` means repeat 0 or more times
    - `[...]` matches any character from between the `[` and the `]`
- `str_c()` combines string arrays, element by element
- `str_detect()` returns `TRUE` or `FALSE` that it detects the pattern in the string
- `str_to_lower()` changes a string to lower case
- `str_to_upper()` changes a string to upper case
- `str_sub()` may also be used for assignment
- See the cheat sheet for many more

  
```{r vowel}
df4 = df_colors %>%
  filter(str_detect(colors, "^[aeiou].*[aeiouy]$"))
df4

df4 = df4 %>%
  mutate(new_colors = str_c(str_to_upper(str_sub(colors,1,1)), # capitalize first letter
                            str_sub(colors,2,-3), #start and end points are inclusive
                            "XX"))

df4
```



> Find all colors names which begin with a vowel, end with a consonant, have no digits in the name, and have between 4 and 10 total characters

```{r}
prob5 = df_colors %>% 
  filter(str_detect(colors, "^[aeiou][:alpha:]{2,8}[^aeiouy0-9]$"))

prob5
```






#### Solution

- Regular expression `[:digit:]` stands for a digit.
    - Alternatives are `\d` or `[0-9]`
- In an R string, need to double the `\` in order to represent the literal `\`.
- A `^` right after `[` negates the matches between the `[` and the `]`
- `{n}` repeats the previous pattern `n` times
- `{m,n}` repeats the previous pattern from `m` to `n` times

```{r}
df5 = df_colors %>%
  filter(!str_detect(colors,"[:digit:]")) %>% # no digit
  filter(str_detect(colors, "^[aeiou].{2,8}[^aeiouy]$")) %>%
  mutate(length = str_length(colors)) # added to check the length
df5
```



> Find all color names that repeat a letter twice in a row.

```{r}
prob6 = df_colors %>% 
  filter(str_detect(colors, "([:alpha:])\\1")) 

prob6
```







#### Solution

- Surrounding a pattern with `()` changes the order of preference and forms a group.
- Whichever pattern matches a group may be referred to by `\x` where `x` is the number of the group
    - For example, `([a-e])\1` will match double letters from a through e.
    - But in a string in R, we need to precede `\` by another `\` to indicate we mean the literal `\`. So, `"([a-e])\\1"`
    
```{r}
## Lecture
df6 = df_colors %>%
  filter(str_detect(colors, "([:alpha:])\\1"))

df6
```

> Find all color names which contain the substrings "red", "blue", or "green"


```{r}
prob7 = df_colors %>% 
  filter(str_detect(colors, "red|blue|green"))

prob7
```






#### Solution

- separate alternative patterns with a `|`

```{r}
## Lecture
df7 = df_colors %>%
  filter(str_detect(colors, "red|blue|green"))

df7
```





