---
title: "Text Analysis with stringr"
author: "Bret Larget"
output: html_document
editor_options: 
  chunk_output_type: console
---

Lecture based on notes by Jessi Kehe

### Setup details

- You will need the packages **tidyverse** and **stringr** for this lecture. (**stringr** should load with **tidyverse**, but if not, run `library(stringr)` separately)

- We will also be using the packages **gutenbergr** and **tidytext** in order to obtain some data so be sure to install them before moving forward.

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,
                      cache=TRUE, autodep=TRUE,
                      cache.comments=FALSE)
library(tidyverse)
library(gutenbergr)
library(tidytext)
## viridis colors
source("../../scripts/viridis.R")
```

## Overview

- We are going to explore a text from **gutenbergr**, which is an R package that provides an interface for downloading public domain texts in the Project Gutenberg collection.  

## Data: gutenbergr

- The library **gutenbergr** contains several hidden data frames with information about a number of texts in the Project Gutenberg collection.
- One is named `gutenberg_metadata`.
- We can access it with the `data()` command

```{r}
data("gutenberg_metadata")
```


```{r data}
## Look at the structure of the gutenberg meta data
str(gutenberg_metadata)

## And column names
names(gutenberg_metadata)
```

- Look at the first several titles

```{r gutenbergr-titles}
gutenberg_metadata %>%
  select(title)
```

- With over 50,000 texts it would be hard to look through to find a text we want.  
- Fortunately we have `dplyr` and `stringr` to help us!

- Let's find a title of a text... How about Huck Finn?  Now what is the full title?
- I can't remember.  Instead, let's do a search for `huck` and see if we can find a match.

```{r huck}
gutenberg_metadata %>%
  filter(str_detect(title, "huck")) %>% 
  select(gutenberg_id, title, author)
```

- Weird, this classic seems to not be in here.  Maybe some letters should be capitalized?

```{r huck-2}
gutenberg_metadata %>%
  select(gutenberg_id, title, author) %>%
  filter(str_detect(title, "(?i)huck" )) # (?i) makes it match case insensitive
```

- There it is!  Let's grab the work using `gutenberg_id = 76`

```{r}
huck = gutenberg_download(76)
head(huck)
```

- We see that the format of the downloaded text is a tibble with two columns.
    - `gutenberg_id`: the same value, 76, for all cases
    - `text`: text of the entire book, including foreward material
- The text of the entire book is broken into multiple lines.

## Louisa May Alcott

- Find the titles of all books written by Louisa May Alcott

```{r get-lw}
alcott = gutenberg_metadata %>%
  select(gutenberg_id, title, author, gutenberg_author_id) %>%
  filter(str_detect(author, "(?i)alcott"))
```

- There are multiple authors with the name Alcott.
- Select those with `author_id == 102`

```{r}
alcott = gutenberg_metadata %>%
  select(gutenberg_id, title, author, gutenberg_author_id) %>%
  filter(gutenberg_author_id == 102)

count(alcott)

alcott %>% 
  distinct(title) %>% 
  count()
```

- There are 50 titles here under her name, but there are some repeats (only 46 unique book titles).

## Little Women

- *Little Women* is a classic in American literature found on many lists of such titles
    - [Here is one!](https://www.goodreads.com/shelf/show/classic-american-literature)
    
```{r}
## We will look at little women
gutenberg_metadata %>%
  filter(str_detect(title, "^(?i)little women$"))

## gutenberg_id = 514...
## The other option does not seem to have a text available
lw_orig = gutenberg_download(514)
```


> Make a data frame with a chapter index, chapter titles, and the text

- We will represent the entire book with one row per chapter:
    - A chapter index with numerical values of the chapters
    - A text variable with the chapter titles
    - A text variable with the entire text of the corresponding chapter
- We will discard the foreword portion of the book.

## First, get the chapter names in a data frame

- By inspection, we see the first 69 lines contains foreword information.
- The chapter titles are in rows 14--66, if we eliminate rows with `text` that begin `PART` or blank lines.

```{r chapters}
## Get chapter names
chapter_names = lw_orig %>%
  select(text) %>%
  slice(14:69) %>%
  filter(!str_detect(text, "PART")) %>% 
  filter(!(text == ""))

chapter_names %>% 
  pull(text)
```

- This is not quite what we wanted....
- The chapter names contain extra spaces at the start
    - It looks like this is done so that the names of the chapters all begin at the same position.
- We can remove these with the `str_trim()` function which removes white space at the start and end of each string.
- The function `str_squish()` also eliminates multiple adjacent spaces.

```{r chapters-2}
## Get chapter names, take 2
chapter_names = lw_orig %>%
  select(text) %>%
  slice(14:69) %>%
  filter(!str_detect(text, "PART")) %>% 
  filter(!(text == "")) %>% 
  mutate(text = str_trim(text))

chapter_names %>% 
  pull(text)
```

- Next, it would be nice to separate each chapter name from its number
- Each number is separated from the chapter name by two spaces
- Use `separate()`
- Then, we can also add a numerical index

```{r}
## chapter names, take 3
chapter_names = lw_orig %>%
  select(text) %>%
  slice(14:69) %>%
  filter(!str_detect(text, "PART")) %>% 
  filter(!(text == "")) %>% 
  mutate(text = str_trim(text)) %>% 
  separate(text, into = c("chapter_number", "chapter_title"), sep = "  ") %>% 
  mutate(chapter_index = row_number()) %>% 
  relocate(chapter_index)

chapter_names %>% 
  print()
```

## Wrangling the text

- After the foreword section, the text uses several rows for each chapter.
    - For example, Chapter 1 begins on row 70 with `CHAPTER ONE`
    - There is a blank line followed by the chapter title.
    - Then another blank line followed by the chapter broken into many rows with each paragraph on one or more rows, broken in odd ways, and blank lines in between.

```{r}
lw_orig %>% 
  slice(70:80)
```

- We will capture all of the text after the foreword into a single very long text string
- Then, we will use the function `str_split()` with the key word "CHAPTER" to break the long text string into a data frame with text from one row per chapter
- Next, a little clean-up and organization.
- Finally, join with the `chapter_names` data frame made earlier.
    
### Remove foreword and collapse
    
```{r}
## Collapse the text into one string
lw = lw_orig %>% 
  slice(-c(1:69)) %>% # remove header content
  select(text) %>%
  filter(!str_detect(text,"^$")) %>% #remove the empty strings
  summarize(full = str_c(text, collapse = " ")) 

dim(lw)
length(lw$full)
str_length(lw$full)
```

### Split the text

```{r}
## Split the text into a string for each chapter
## Add a chapter index
lw_text = tibble(
  text = str_split(lw, "CHAPTER ") %>%
    unlist()) %>% 
  ## remove the extra first row
  slice(-1) %>% 
  ## add the index
  mutate(chapter_index = row_number()) %>% 
  relocate(chapter_index)

dim(lw_text)

## # of characters per chapter
lw_text %>% 
  mutate(n = str_length(text)) %>% 
  pull(n)
```

#### Join

```{r}
lw = chapter_names %>% 
  left_join(lw_text)
```

#### Clean up

- Each chapter text begins with:
    - the all-caps chapter number
    - followed by the all-caps chapter title
- As we have the number and title, we can count characters and remove the correct number from each title

```{r}
## Add 3 to account for the spaces
lw = lw %>% 
  mutate(text = str_sub(text,
                        start = str_length(chapter_number) +
                          str_length(chapter_title) + 3,
                        end = -1))
```

## Analysis

> Get the number of times the main characters' names appear by chapter

- The main characters in Little Women are the sisters Amy, Beth, Jo, and Meg, and the neighbor boy Laurie.

```{r chapters-plot}
main = lw %>%
  mutate(Amy = str_count(text,"Amy")) %>%
  mutate(Beth = str_count(text,"Beth")) %>%
  mutate(Jo = str_count(text,"Jo")) %>%
  mutate(Laurie = str_count(text,"Laurie")) %>%
  mutate(Meg = str_count(text,"Meg")) 
```

- Next we will create a plot that displays the cumulative number of times the main characters are mentioned by chapter.
- In order to more easily color lines according to the characters, we will:
    - put the `main` data frame in long form.
    - add a cumulative count variable

```{r}
main_long = main %>%
  pivot_longer(cols = Amy:Meg,
               names_to = "character",
               values_to="count") %>% 
  group_by(character) %>% 
  mutate(cumsum = cumsum(count)) %>% 
  ungroup()
```


```{r}
ggplot(main_long, aes(x = chapter_index, y = cumsum, color = character)) +
  geom_line()+
  xlab("Chapter number") +
  ylab("Cumulative count")+
  labs(color = "Character") +
  ggtitle("Little Women Main Characters")
```

### Interpretations

- Can this graphic help us to understand anything about the plot?

- A few observations:
    - Jo is the principal character and her name appears regularly throughout the chapters
    - With a careful eye, we see that the lines for Amy and Laurie are nearly parallel toward the end of the book, suggesting that they are mentioned together most of the time
        - In fact, they get married to each other
    - Meg's curve is seldom mentioned after about chapter 28
        - In fact, she get married to someone else and drops out of the story

### Another plot

```{r}
ggplot(main_long, aes(x = character, y = count, fill = character)) +
  geom_col() +
  facet_wrap( ~ chapter_index) +
  xlab("") +
  ylab("Count") +
  ggtitle("Little Women Main Characters",
          subtitle = "Counts by Chapter") +
  scale_x_discrete(labels = NULL)
```












