---
title: "Boston Marathon Analysis"
author: "Bret Larget"
output: html_document
---

This R Markdown document includes contributions by Professor Jessi Kehe.

\renewcommand{\prob}{\mathsf{P}}
\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}

### Setup details

- This lecture use the following scripts, assumed to be in your course scripts directory.
    - `COURSE/scripts/viridis.R`
    - `COURSE/scripts/ggprob.R`
    
- You also need the Boston Marathon times
    - `COURSE/data/TIM1.txt`

- You will need the package `tidyverse` for these lectures. 

- The code uses **readr** function `read_table()` which assumes a version number at least 2.0.0.
    - If your **readr** version number is older, please update.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning = FALSE,
                      cache=TRUE, autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(scales)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

## The Boston Marathon

- The Boston Marathon is the world's oldest annual marathon competition.
- It has been run every year since 1897 (one year after the first modern marathon competition in the 1896 Olympic Games), except for 2020 due to the COVID-19 pandemic
- In most years, the Boston Marathon is held on Patriot Day, the third Monday of April, but in 2021 was delayed to October.
- The 2013 race was stopped prematurely, preventing many competitors from finishing, due to a terrorist bombing near the finish line which killed three spectators and injured hundreds of people.
- The length of a marathon race is 26.2 miles, or about 42.65 km.
- Entrance in the Boston Marathon requires obtaining a qualifying time based on age and sex in a qualified marathon race in the 18 months prior.

### Populations and Samples

- The data we use are from all finishers in the years 2010 and 2011.
- The original data includes some results from 2013.
- The purpose the data was collected was to estimate the finishing times of competitors who were not allowed to complete the race due to the bombing.
- Not all individuals who are eligible to compete in the Boston Marathon in a given year actually compete.
- We will consider the observed data as representative samples from the populations of people worldwide eligible to race in the Boston Marathon and would finish the race in the given years.
- Various questions will consider sub-populations based on the age and sex of the competitors.
- The age of a competitor is determined by their age on the date of the race.
- There are different records and sub-competitions based on sex and various age ranges: 18-34, 35-39, 40-44, \ldots, 75-79, and 80 and older.
- There are typically thousands of runners in the race.
- Not all runners can start at the same time: runners begin the race in stages and run with chips in their shoes which allows accurate individual times to be calculated.
- The Madison Marathon was held yesterday, November 14, 2021.

### Data Wrangling

- Data source: [Boston Maraton Data](https://rls.sites.oasis.unc.edu/boston.html)

- The data set `TIM.txt` contains all times from 2010, 2011, and 2013.

- We make the following changes to the original data:
    - Eliminate the year 2013 as many racers did not finish due to the bombing
    - Create a `Sex` variable
    - Sum the times for racers for different race segments to obtain a total time
    - Add an `Age_Range` variable
- Note the use of `read_table()` to read in the data
    - The `col_types` argument is set so that each column is read in as type *double* by default.


```{r}
bm_orig = read_table("../../data/TIM.txt",
                col_types = cols(.default = col_double()))

bm = bm_orig %>% 
  filter(!is.na(`K40-Fin`)) %>% 
  filter(Year < 2013) %>% 
  arrange(Year, BibNum) %>% 
  select(-starts_with("Start"), -HalfMar, -Age2014) %>% 
  mutate(Sex = case_when(
    Gender1F2M == 1 ~ "female",
    Gender1F2M == 2 ~ "male")) %>% 
  mutate(Time = pmap_dbl(select(., starts_with("K")), sum)) %>% 
  mutate(age1 = case_when(
    Age < 35 ~ 18,
    Age >= 80 ~ 80,
    TRUE ~ floor(Age/5)*5),
    age2 = case_when(
      Age < 35 ~ 34,
      TRUE ~ age1 + 4),
    Age_Range = case_when(
      age1 == 80 ~ "80 and older",
      TRUE ~ str_c(age1,"-",age2))) %>% 
  select(-Gender1F2M, -age1, -age2) %>% 
  relocate(BibNum, Year, Sex, Age, Age_Range, Time)

```

### Variables

- The transformed data set has these variables:
    - `BibNum`: bib number, a unique identifier for each competitor within a year
    - `Year`" either 2010 or 2011
    - `Sex`: female or male
    - `Age`: age in years
    - `Age_Range`: from 18-34, 35-39, ..., 75-79, and 80 and older
    - `Time`: total time to finish in decimal minutes
    - `K0-5` through `K40-Fin`: times in decimal minutes for 5 km segments of the race plus the final shorter segment.
    
### Data Explorations

- We present a few data explorations

#### Number of Competitors by Year, Age, and Age Range

```{r}
bm %>% 
  count(Year, Sex, Age_Range) %>% 
  pivot_wider(names_from = Age_Range,
              values_from = n)
```


```{r}
bm %>% 
  filter(Year == 2010) %>% 
ggplot(aes(x = Sex, fill = Sex)) +
  geom_bar() +
  xlab("") +
  ylab("# of Finishers") +
  ggtitle("2010 Boston Marathon Finishers") +
  facet_wrap(~Age_Range)

bm %>% 
  filter(Year == 2011) %>% 
ggplot(aes(x = Sex, fill = Sex)) +
  geom_bar() +
  xlab("") +
  ylab("# of Finishers") +
  ggtitle("2011 Boston Marathon Finishers") +
  facet_wrap(~Age_Range)
```

#### Times

```{r}
bm %>% 
  group_by(Year, Sex, Age_Range) %>% 
  summarize(n = n(),
            min = min(Time),
            q10 = quantile(Time, 0.10),
            median = median(Time),
            q90 = quantile(Time, 0.90)) %>% 
  print(n = Inf)
```


```{r}
bm %>% 
  filter(Year == 2010) %>% 
ggplot(aes(x = Sex, y = Time, fill = Sex)) +
  geom_boxplot(coef = Inf) +
  xlab("") +
  ylab("Finish Times") +
  ggtitle("2010 Boston Marathon Finishers") +
  facet_wrap(~Age_Range)

bm %>% 
  filter(Year == 2011) %>% 
ggplot(aes(x = Sex, y = Time, fill = Sex)) +
  geom_boxplot(coef = Inf) +
  xlab("") +
  ylab("Finish Times") +
  ggtitle("2011 Boston Marathon Finishers") +
  facet_wrap(~Age_Range)
```

## Estimation of a Population Mean

- Let's consider females marathon runners good enough to qualify for and finish the 2010 Boston Marathon, aged 18-34, and estimate the mean time such runners would have finished had they competed.
- We consider the population to be all women aged 18-34 who are eligible for the Boston Marathon and able to finish to complete the course in 2010.

### Statisical Model

- Let $\mu$ be the mean time in minutes for this population.
- Our data is the 3557 times of the women in this age group who actually did complete the race, $x_1, \ldots, x_n$ for $n = 3557$.
- We model the data as if being drawn from some distribution $F$ with mean $\mu$ and standard deviation $\sigma$ and as if they are independent observations.

$$
X_i \sim F(\mu, \sigma) \quad \text{for $i = 1, \ldots, n$}
$$

### Data Summary

- The finishing times for our sample.

```{r}
prob1 = bm %>% 
  filter(Sex == "female",
         Age_Range == "18-34",
         Year == 2010)

prob1_sum = prob1 %>% 
  summarize(n = n(),
            mean = mean(Time),
            sd = sd(Time))

prob1_sum

ggplot(prob1, aes(x = Time)) +
  geom_density(fill = "lightpink", color = "black") +
  geom_hline(yintercept = 0) +
  xlab("Finishing Time") +
  ggtitle("2010 Boston Marathon",
          subtitle = "Women aged 18-34") +
  theme_minimal()
  
```

- Our point estimate is $\bar{x} = `r round(prob1_sum$mean, 3)`$ minutes
- Note that the distribution of finishing times in the same is not normally distributed: will this affect our method to construct a confidence interval?
- We are interested in the **sampling distribution of the sample mean**


### Simulation

- To explore the normality question, we do the following simulation experiment
- Treat the sample as an estimate of the underlying population
- Take random samples of size 3557 (with replacement) from the sample
- Calculate the sample mean of each resampled data set
- Repeat many times
- Examine center (mean), spread (standard deviation), and shape of the sampling distribution of the sample mean

```{r}
B = 50000
x = prob1 %>% pull(Time)

set.seed(20211115)
sample_means = tibble(
  xbar = map_dbl(1:B, ~{
  return( mean(sample(x, replace = TRUE)) )
}))

sim1_sum = sample_means %>% 
  summarize(n = n(),
            mean = mean(xbar),
            sd = sd(xbar))

sim1_sum

ggplot(sample_means, aes(x = xbar)) +
  geom_density(fill = "lightpink", color = "black") +
  geom_norm_density(sim1_sum$mean, sim1_sum$sd) +
  geom_hline(yintercept = 0) +
  xlab("Finishing Time Sample Means") +
  ggtitle("2010 Boston Marathon, Women aged 18-34",
          subtitle = "Sampling Distribution of xbar, n = 3557") +
  theme_minimal()
```

- The ideal blue normal density matching the mean and sd of the sampling distribution is very close to the simulated density

### Repeat for Smaller n

- Let's look at the shape if $n=50$ instead

```{r}
B = 50000
n = 50
x = prob1 %>% pull(Time)

set.seed(20211115)
sample_means = tibble(
  xbar = map_dbl(1:B, ~{
  return( mean(sample(x, size = n, replace = TRUE)) )
}))

sim2_sum = sample_means %>% 
  summarize(n = n(),
            mean = mean(xbar),
            sd = sd(xbar))

sim2_sum

ggplot(sample_means, aes(x = xbar)) +
  geom_density(fill = "lightpink", color = "black") +
  geom_norm_density(sim2_sum$mean, sim2_sum$sd) +
  geom_hline(yintercept = 0) +
  xlab("Finishing Time Sample Means") +
  ggtitle("2010 Boston Marathon, Women aged 18-34",
          subtitle = "Sampling Distribution of xbar, n = 50") +
  theme_minimal()
```

- Note that the sampling distribution of the sample mean has an approximate normal shape, even for a much smaller sample size such as $n=50$, despite the skewed population.

### Back to a Confidence Interval

- The sample mean is $\bar{x} = `r round(prob1_sum %>% pull(mean), 3)`$
- The sample standard deviation of finishing times is
$s = `r round(prob1_sum %>% pull(sd), 3)`$
- The SE estimated from the simulation is
$\text{SE}(\bar{x}) = `r round(sim1_sum %>% pull(sd), 3)`$

$$
(\text{point estimate}) \pm z \text{SE}
$$

$$
235.505 \pm 1.96(0.609)
$$

```{r}
z = qnorm(0.975)
ci = prob1_sum$mean + c(-1,1)*z*sim1_sum$sd
round(ci,3)
```

### Interpretation

> We are 95% confident that the mean finishing time of women aged 18-34 capable of finishing the Boston Marathon in 2010 would have been between 234.31 and 236.70 minutes.

#### Rounding Rule of Thumb

- A good rule of thumb for rounding is to round the margin of error to two significant digits and round the estimate to the same accuracy.

## Formula Instead of Simulation

- As with proportions, there is a formula we could have used instead of simulation to determine the standard error

$$
\text{SE}(\bar{x}) = \frac{\sigma}{\sqrt{n}}
$$

- where $\sigma$ is the population standard deviation.

- We can replace the $\sigma$ with the sample standard deviation.
    - When $n$ is quite large, such as $n = 3557$ in our example, $s$ is typically a very accurate estimate of $\sigma$
    - Examine our example
    
```{r}
## Simulation SE
sim1_sum$sd

## Formula
prob1_sum$sd / sqrt(prob1_sum$n)
```
    
## Built-in R Function

- There is also a built-in R function `t.test()` which does a hypothesis test and a confidence interval.
- Here, ignore the hypothesis test output
- Note the use of the t distribution which accounts for replacing $\sigma$ with $s$
    - For $n = 3557$, this makes very little difference.
    
```{r}
t.test(x)
```

- Instead of using the 0.975 quantile from a standard normal distribution, $z = `r qnorm(0.975)`$
- Use instead $t = `r qt(0.975, 3557-1)`$

## More Problems

- If time, do confidence intervals for other subsets of the data

#### Smaller sample size

- Here is an artificial example with $n=50$
- Use a quantile from the t distribution instead of standard normal

```{r}
## new sample
n = 50
new_x = sample(x, size = n)
new_x_mean = mean(new_x)
new_x_sd = sd(new_x)
new_x_se = new_x_sd / sqrt(n)
t_mult = qt(0.975, n-1)
new_x_mean
new_x_sd
new_x_se
t_mult

new_ci = new_x_mean + c(-1,1) * t_mult * new_x_se
new_ci

## compare to t.test
t.test(new_x)
```

> For the artificial sample of size $n=50$, we are 95% confident that the mean finishing time among all women aged 18-34 who completed the Boston Marathon is between 225.3 and 246.6 minutes.

- Note that this confidence interval did capture the sample mean from all 3557 women.
- We would expect about 95% of confidence intervals from artificial samples of size $n=50$ to include the actual sample mean.

### Explore 95% Capture Probability

- Repeatedly take samples of size 50 from the sample
- Calculate 95% confidence intervals
- See how many contain the sample mean from all 3557

```{r}
B = 50000
n = 50
t_mult = qt(0.975, n-1)
x = prob1$Time
mu = mean(x)
sigma = sd(x) * sqrt(3556/3557)
mu
sigma

capture_base = vector("list", B)

capture_sim = map_dfr(capture_base, ~{
  samp = sample(x, size = n, replace = TRUE)
  out = tibble(xbar = mean(samp),
         sd = sd(samp),
         se = sd/sqrt(n),
         low = xbar - t_mult*se,
         high = xbar + t_mult*se)
  return ( out )
})
```

### Find captures

```{r}
capture_sim = capture_sim %>% 
  mutate(capture = low < mu & mu < high) %>% 
  mutate(index = row_number()) %>% 
  relocate(index)

capture_sim %>% 
  summarize(mean(capture))

capture_sim %>% 
  sample_n(100) %>% 
  mutate(index = row_number()) %>% 
ggplot(aes(x = low, xend = high, y = index, yend = index, color = capture)) +
  geom_segment() +
  geom_vline(xintercept = mu, color = "red", linetype = "dashed") +
  xlab("Mean Marathon Time") +
  ggtitle("2010 Boston Marathon, Women aged 18-34",
          subtitle = "Sampling Distribution of xbar, n = 50") +
  theme_minimal()
  
```



