---
title: "STAT 240 Discussion 12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
library(tidymodels)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

## Add names of all group members who participate

ADD GROUP NAMES HERE


## Questions

## Data

The following code reads in the official Madison weather data
and calculates the average winter temperature (November through February) for each winter and graphs the data and a fitted linear model.

The official weather record in this data set extends from January 1, 1869 through December 31, 2020.
The complete winters begin with 1869-1870 in November and extend through the winter of 2019-2020 as this data set does not include the most recent winter.

```{r read-data}
## Read and transform the Madison weather data
mw_orig = read_csv("../../data/madison-weather-official-2020.csv")

mw = mw_orig %>% 
  mutate(year = year(date),
         month = month(date, label = TRUE),
         .after = date) %>% 
  mutate(year1 = case_when(
    month > "Jun" ~ year,
    month < "Jul" ~ year - 1),
    year2 = year1 + 1,
    winter = str_c(year1, "-", year2),
    .before = year) 

## add the winter variables and filter
mw_winter = mw %>%
  filter(month %in% c("Nov", "Dec", "Jan", "Feb")) %>%
  select(-prcp, -contains("snow")) %>% 
  drop_na() %>% 
  filter(winter >= "1869-1870" & winter <= "2019-2020") %>% 
  group_by(winter, year1) %>%
  summarize(tavg = mean(tavg))
```

### Scatter plot and regression line

```{r plot-data}
ggplot(mw_winter, aes(x = year1, y = tavg)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  xlab("Year (first year of the winter)") +
  ylab("Average Winter Temperature") +
  ggtitle("Madison, WI Average Winter Temperature",
          subtitle = "Winter defined as Nov. through Feb.") +
  theme_bw()
```

## Regression Model

The next block of code fits a linear model using the **tidymodels** functions and then uses the `augment()` function from **broom** to add columns `.resid` and `.pred` to the data set `mw_winter` with the residuals and predicted values.

Note this use differs from the earlier approach we used from **modelr** and the functions `add_predictions()` and `add_residuals()`.

```{r fit-regression-model}
## tidymodels fit of the regression model
winter_lm = linear_reg() %>% 
  set_engine("lm") %>% 
  fit(tavg ~ year1, data = mw_winter)

## extracting a summary table of coefficients and inference values
winter_lm_sum = tidy(winter_lm)

winter_lm_sum

## a one-line summary of the fitted model
winter_lm_glance = glance(winter_lm) %>% 
  print(width = Inf)

winter_lm_glance %>% 
  print(width = Inf)

## augment residuals and predicted values
mw_winter_aug = winter_lm %>% 
  augment(mw_winter)

head(mw_winter_aug)
```


## Problems


### 1 

Using the results of the linear model fit above, write out the estimated linear model by replacing the (a) and (b) with the estimated coefficients; y_hat represents the estimated average winter temperature in Madison for year x.  What is the interpretation of the estimated slope?

```{r}

```


y_hat = (23.0) + () * x

As the esitimated slope is positive, it indicates that the average Madison winter temperature has been rising




### 2A

Calculate the end points of a 95% confidence interval for the slope of the regressions model.
Note that we have data for this entire period of time,
so a confidence interval here represents the slope in a linear unseen climate trend whereas the data is affected by the randomness of weather.

Round the margin of error to two significant digits and then round the estimated slope to the same accuracy.
The margin of error is the product of the multiple of a quantile from a t distribution with an appropriate number of degrees of freedom.

```{r}

```

### 2B

Interpret this confidence interval in context, following examples from lecture.


REPLACE THIS TEXT WITH YOUR RESPONSE

### 2C

What is the point estimate and end points of a 95% confidence interval for the total change in average winter temperature in Madison over the past 150 years?




### 3

Is there strong evidence that the unknown slope of the regression line of average winter temperature in Madison due to a changing climate is different from zero?

### 3A
Report a t test statistic and p-value from the model summary output.

REPLACE THIS TEXT WITH YOUR RESPONSE

### 3B

Verify the calculation of the p-value using the `pt()` function.

```{r}

```

### 3C

Write a summary of this test in context.

REPLACE THIS TEXT WITH YOUR RESPONSE







### 4

Make a density plot of the residuals from the model. Calculate the mean and standard deviation of the residuals. Overlay a normal density with these values for the mean and standard deviation.

Does it appear that a normal distribution is a reasonable approximation of the distribution of variation of points around the regression line?

```{r}

```

REPLACE THIS TEXT WITH YOUR RESPONSE




### 5

Make a scatter plot with `year1` on the x axis and the residuals on the y axis. Add a horizontal line with a y intercept of zero.

Are there strong patterns in the residual plot, or do the residuals resemble random noise? Briefly explain.

```{r}

```

REPLACE THIS TEXT WITH YOUR RESPONSE










