---
title: "STAT 240 Discussion 9"
output: html_document
---

Group 311C: William Fan, Jason Fu, Fei Qin, Nihar Simhadri, **Kai Tsuyoshi** 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE,
                      cache=TRUE, autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(scales)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}

## Problems

The questions in this discussion assignment use the chimpanzee data set.
In answering these questions, you may assume the following in models for the data:

1. all trials are independent
2. trials made with a partner have the same chance of being pro-social, even if the identities of the actors or partners differ.

### Problem 1

#### A

Using the chimpanzee data,
combine all of the trials made without a partner.
Count the number of pro-social and selfish choices made.

```{r}
chimpanzee = read_csv("../../data/chimpanzee.csv") 

no_part = chimpanzee %>%
  filter(partner == "none") %>% 
  summarize(prosocial = sum(prosocial),selfish = sum(selfish), )

no_part
```

#### B

Write a model for the number of pro-social choices.

-$X$ is number of pro-social choices made by chimpanzees without partners in 180 trials 
- Assume:
    - each trial could be pro-social or selfish
    - 90 trials was predetermined
    - trials are independent (this assumption is debatable)
    - same probability of pro-social choice for each trial (this assumption is debatable)
- Under these assumptions, $X$ has a binomial distribution. 
**(taken from lecture notes)**

-$n = 180$ => number of trials
-$p$ is the probability that the chimpanzees make the pro-social choice
-$x = 83$ => number of times the pro-social choice was made in the above dataset for chimpanzees without partners and thus
    -$\hat{p} = \frac{x}{n} = \frac{83}{180} = `r round(83/180,3)`$

$$
X \mid p \sim \text{Binomial}(90,p)
$$


#### C

Use the Agresti-Coull method to construct a 95% confidence interval for the probability that a chimpanzee makes the pro-social choice in experimental conditions when there is no partner present.

$$
\tilde{p} \pm 1.96 \sqrt{ \frac{\tilde{p}(1-\tilde{p})}{n+4} }
$$

```{r}
binom_se =  function(n, p)
{
  return ( sqrt( p*(1-p)/n) )
}

binom_ci = function(est, se, conf=0.95)
{
  z = qnorm(1 - (1 - conf)/2)
  me = z * se
  ci = est + c(-1,1)*me
  return(ci)
}


x = 83
n = 180

p_tilde = (x+2)/(n+4)
p_tilde

se_agresti = binom_se(n+4, p_tilde)
se_agresti

ci_agresti = binom_ci(p_tilde, se_agresti)
round(ci_agresti, 3)
```

#### D

Interpret this confidence interval in the context of the experimental setting.

We are 95% confident that the probability of a chimpanzee without a partner making a pro-social choice is between 0.390 (39%) and 0.534 (54.3%)

### Problem 2

Using the same data as in the previous problem,
do the following.

#### A

State hypotheses for a two-sided hypothesis test that the probability that a chimpanzee makes the pro-social choice over the long run is equal to 0.5 versus the alternative that this probability is not 0.5.

- $H_0: p = 0.5$
- $H_a: p \neq 0.5$
    

#### B

Pick a test statistic, state the sampling distribution of this statistic if the null hypothesis is true,
and compute a p-value for the test.

```{r}

gbinom(180, 0.5, scale=TRUE) + 
  geom_vline(xintercept = 83, color="red",
             linetype = "dashed") +
  
  theme_bw()

 p_val = sum(dbinom(c(0:83, 97:180), 180, 0.5))
 p_val
```

#### C

Interpret these results in the context of the experiment following the examples from lecture notes. 

There is strong evidence that the null hypothesis is true, as running a left hand or right hand would both indicate a very common occurrence of such a result, and that $\hat{p}$ is close to 0.5

### 3

Use functions from the file `ggprob.R` to visualize the calculation of the p-value by graphing an appropriate probability distribution and using differential coloring to indicate what in the plot corresponds to the p-value.
Include a title and meaningful axis labels.

```{r}
# Left-hand
gbinom(180, 0.5, a = 0, b = 83) +
  geom_binom_density(180, 0.5, a =84, b = 85, color = "gray") +
  geom_hline(yintercept = dbinom(83, 180, 0.5), color = "red", linetype = "dashed") +
  theme_bw()

```

```{r}
# Right-hand
gbinom(180, 0.5, a = 97, b = 180) +
  geom_binom_density(180, 0.5, a = 95, b = 96, color = "gray") +
  geom_hline(yintercept = dbinom(83, 180, 0.5), color = "red", linetype = "dashed") +
  theme_bw()
```

